---
title: "Assignment 3"
author: "Jordi Florit Ensenyat, Josep Gabriel Fornés Reynés, Pau Girón Rodríguez"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **1. Introduction**

This assignment aims to build predictive models to classify the loan
status variable (approved/not approved) for potential applicants using
the loan_data.csv dataset. The dataset, available on Kaggle, is a
synthetic version inspired by the original Credit Risk dataset and
enriched with additional variables based on Financial Risk for Loan
Approval data. The analysis will be structured as follows:

-   **Data Analysis**: Analyze key features, distribution patterns,
    relationships, plots, missing data, and outliers to better
    understand credit risk factors.

-   **Data Processing**: Prepare the data by handling missing values,
    outliers, and normalizing features to ensure the dataset is ready
    for modeling.

-   **Classification Methods**: Evaluate different classification
    methods for modeling loan approval, using various structures and
    parameters, and compare the models obtained.

This structured approach ensures a comprehensive analysis and comparison
of different models to identify the most effective one for predicting
loan approval.

### **1.1. Dataset Information**

The dataset, available on Kaggle
(<https://www.kaggle.com/datasets/taweilo/loan-approval-classification-data>)
is a synthetic version inspired by the original Credit Risk dataset and
enriched with additional variables based on Financial Risk for Loan
Approval data. SMOTENC was used to simulate new data points to enlarge
the instances. The dataset is structured for both categorical and
continuous features.

### **1.2. Variable Information**

The dataset contains 45,000 records and 14 variables. Below is a
detailed description of each variable:

***1. person age***: Age of the person (numeric)

***2. person gender***: Gender of the person (categorical: female, male)

***3. person education:*** Highest education level (categorical:
Associate, Bachelor, Doctorate, High School, Master)

***4. person income:*** Annual income (numeric)

***5. person emp exp:*** Years of employment experience (integer)

***6. person home ownership:*** Home ownership status (categorical:
MORTGAGE, OTHER, OWN, RENT)

***7. loan amnt:*** Loan amount requested (numeric)

***8. loan intent:*** Purpose of the loan (categorical:
DEBTCONSOLIDATION, EDUCATION, HOMEIMPROVEMENT, MEDICAL, PERSONAL,
VENTURE)

***9. loan int rate:*** Loan interest rate (numeric)

***10. loan percent income:*** Loan amount as a percentage of annual
income (numeric)

***11. cb person cred hist length:*** Length of credit history in years
(numeric)

***12. credit score:*** Credit score of the person (integer)

***13. previous loan defaults on file:*** Indicator of previous loan
defaults (categorical: No, Yes)

#### **Class label (target variable)**

***14. loan status:*** Loan approval status (integer 1 = approved; 0 =
rejected)

### **1.3. Research Question**

Utilize the data provided to build predictive models to classify the
loan status variable (approved/not approved) for potential applicants.
To answer this question you will:

***a) (5 POINTS) Perform Explanatory Data Analysis:*** analyze key
features, distribution patterns, relationships, plots, missing data,
outliers, ... to better understand credit risk factors.

***b) (20 POINTS) Consider different classification methods for
modelling loan approval.*** For each method, use different structures
and/or parameters, different training/testing sets, compare the
different models obtained and comment your results. Which model is the
“best”?

## **2. Data Analisis**

### **2.1. Importing Data**

First of all, we have to import our data. Here, we are loading the
necessary libraries and reading the CSV file containing the loan data.

```{r}
# Load the necessary libraries and read the CSV file containing the loan data
library(tidyverse)
library(ggcorrplot)
library(gridExtra)
library(ggplot2)
library(dplyr)
library(e1071)
library(caret)
library(gmodels)
library(rpart)
library(rpart.plot)



data <- read.table("loan_data.csv", header = TRUE, sep = ",")
```

This is one simple example of the dataset:

```{r}
# Display the structure of the dataset
str(data)
```

The summary data. The summary() function gives a statistical
summary of each column, including minimum, maximum, mean, and quartiles.

```{r}
# Display a statistical summary of each column, including minimum, maximum, mean, and quartiles
summary(data)
```

### **2.2. Missing Values**

```{r}
# Identify missing values
colSums(is.na(data))
```

The function counts the number of missing values in each column of the
dataframe. We can see that there are no missing values. This confirms
that there are no NA values in the dataset, which is crucial to avoid
errors in further analysis.

### **2.3. Outliers**

As we can see into the summary, there are some **unusual data
(outliers)**:

***· person_age***: the maximum age of a person in this column is 144,
which is not very credible and strange data. So now, we are going to
show the possible outliers into the person_age data. Since we are
talking in terms of requesting loans, we are going to assume that people
over 80 years old are not going to request a loan. Now let's visualize
people over 80 years old.

```{r}
# Filter rows directly to identify potential outliers in person_age
outliers_age <- data[data$person_age > 80, ]

# Display the first few rows
outliers_age
```

***· person_emp_exp***: there is a person with 125 years of work
experience, which makes us think that it is unusual and erroneous data.
Given that work experience generally begins between 15 and 25 years of
age, and considering an average retirement around 65 years of age, a
reasonable range for work experience could be 0 to 50 years. Now let's
look at the people who are in a higher rank.

```{r}
# Filtrar filas directamente
outliers_emp_exp <- data[data$person_emp_exp > 50, ]

# Mostrar las primeras filas
outliers_emp_exp
```

```{r}
# Boxplot for person_age
boxplot(data$person_age,
        main = "Boxplot of Age (person_age)",
        xlab = "Age",
        horizontal = TRUE,
        col = "skyblue")

# Boxplot for person_emp_exp
boxplot(data$person_emp_exp,
        main = "Boxplot of Work Experience (person_emp_exp)",
        xlab = "Years of Experience",
        horizontal = TRUE,
        col = "lightgreen")
```

After analyzing the previous tables, we have drawn the following
conclusions regarding the identified outliers:

**1. Overlap between age and work experience:**

· There are 9 individuals who have both more than 80 years of age and
more than 50 years of work experience.

**2. Outlier classification among these 9 individuals:**

· Out of these 9 individuals, 7 are considered outliers, as their age
and work experience values are unusually high and implausible.

· Among the remaining two individuals:

· The person with 84 years old and 61 years of work experience seems
reasonable within the context and will not be considered an outlier.

· The person with 94 years old and 76 years of work experience appears
less credible, although theoretically possible. However, we decided to
classify this individual as an outlier due to its high improbability.

**3. Individuals with extreme values individually (age or work
experience):**

· Those with less than 80 years of age or less than 50 years of work
experience are not considered outliers. While their data may seem
unusual, they will remain in the analysis as potentially valid entries.

**4. Impact on modeling:**

· The values identified as non-outliers will be included in the
predictive models, while extreme values classified as outliers will be
removed to avoid biases in the analysis.

```{r}
# Filter values considered as non-outliers
data_cleaned <- data %>%
  filter(
    # Keep those with work experience <= 50 or plausible exceptional cases
    !(person_age >= 94 & person_emp_exp >= 76), # Case considered outlier
  )

# Verify the cleaned data
summary(data_cleaned)

# Save the cleaned data
write.csv(data_cleaned, "loan_data_cleaned.csv", row.names = FALSE)
```

With this new data, we can see that there is a person whose
***person_income is 2448661***. This is a very high value, but we can
consider it valid, because it is possible.

### **2.4. Relationships**

Now we are going to identify **highly correlated variables**. This is
important because highly correlated variables can cause problems in
certain models. Correlation can only be calculated between numeric
variables. Values range from -1 (perfect negative correlation) to 1
(perfect positive correlation). Values close to 0 indicate little or no
variation. It is not desirable to have correlated features if we are
modeling the data using linear regressions.

```{r}
# Calculate correlations between numeric variables
numeric_data <- data_cleaned %>% select_if(is.numeric)
corr_matrix <- cor(numeric_data, use = "complete.obs")

# Plot the correlation heatmap
ggcorrplot(corr_matrix, method = "square", type = "full", lab = TRUE, lab_size = 3, 
           colors = c("#6D9EC1", "white", "#E46726"), title = "Correlation Heatmap")
```

We are going to carry out the study of the previous correlations.

-   **Stronger correlations**:

    -   **cb_person_cred_hist_length and person_age**: correlation of
        0.88, which suggests that the older the person is, the longer
        their credit history is. This is to be expected, since older
        people tend to have a longer financial history.

    -   **person_age and person_emp_exp**: correlation of 0.95 is an
        extremely strong correlation between age and work experience.
        This makes sense, since generally, the older you are, the more
        time you have spent working.

    -   **person_emp_exp and cb_person_cred_hist_length**: work
        experience is also highly related to credit history (0.84). This
        reinforces the idea that these three variables (person_age,
        person_emp_exp, cb_person_cred_hist_length) contain redundant
        information, since they all represent, in some way, time in the
        work or financial system.

This strong correlation suggests that these variables are related, and
it may be redundant to include both in a predictive model. That can
cause multicollinearity problems. It is recommended to delete one of
them or transform them.

As this separation is only recommended for linear regression models, we
are not going to perform the separation.

-   **Moderate correaltions**:

    -   **loan_amnt and loan_percent_income**: correlation of 0.59, this
        implies that people who request larger loans tend to dedicate a
        greater percentage of their income to repaying them.

They appear to be important variables, as they have moderate
correlations with loan status (loan_status) and with each other. These
variables are relevant to predict the status of the loan.

-   **loan_status and loan_percent_income**: correlation of 0.38,
    showing that loans are approved more easily if the percentage is not
    excessively high.

-   **loan_status and loan_int_rate**: correlation of 0.33, this may
    indicate that approved loans tend to be associated with specific
    interest rates, perhaps reflecting that rates are aligned with
    accepted risk levels.

Weak correlations:

-   **loan_amnt and person_income**: there is a weak but positive
    correlation (0.31) between loan amount and income. This suggests
    that income is a relevant factor, but not a determining factor for
    the amount requested.

-   **person_income and cb_person_cred_hist_length**: correlation of
    0.13, it could be because having a long credit history is not
    necessarily linked to high income, but rather to access to credit.

Weak or non-existent correlations:

-   **loan_status and other variables (except credit_score)**: the
    correlation with loan_status is weak with most variables, implying
    that there is no strong linear relationship between them.

-   **credit_score with other variables**: has very weak correlations
    with most variables, such as loan_amnt (0.01) and loan_int_rate
    (0.01). This could indicate that credit score is not directly
    related to these variables or that the relationship is not linear.
    Although this variable is important in the financial world, its low
    correlation with other variables could imply that it does not
    provide much predictive value in this dataset (or that the
    relationship is non-linear).

Now we are going to analyze the variables to find out if they have
little or no variance:

```{r}
# Identify variables with low variance
sapply(data_cleaned, function(x) length(unique(x)))
```

Most variables with low variance are categorical variables. Let's see
representations of these variables:

```{r}
# Bar plot for gender distribution
plot_gender <- ggplot(data_cleaned, aes(x = person_gender)) +
  geom_bar(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(
    title = "Gender Distribution",
    x = "Gender",
    y = "Frequency"
  ) +
  theme_minimal()

# Bar plot for home ownership status
plot_home <- ggplot(data_cleaned, aes(x = person_home_ownership)) +
  geom_bar(fill = "lightgreen", color = "black", alpha = 0.7) +
  labs(
    title = "Home Ownership Distribution",
    x = "Home Ownership Type",
    y = "Frequency"
  ) +
  theme_minimal()

# Bar plot for education level
plot_education <- ggplot(data_cleaned, aes(x = person_education)) +
  geom_bar(fill = "lightpink", color = "black", alpha = 0.7) +
  labs(
    title = "Education Level Distribution",
    x = "Education Level",
    y = "Frequency"
  ) +
  theme_minimal()

# Bar plot for loan intent
plot_intent <- ggplot(data_cleaned, aes(x = loan_intent)) +
  geom_bar(fill = "coral", color = "black", alpha = 0.7) +
  labs(
    title = "Loan Intent Distribution",
    x = "Loan Intent",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 10, hjust = 1))

# Bar plot for previous loan defaults
plot_defaults <- ggplot(data_cleaned, aes(x = previous_loan_defaults_on_file)) +
  geom_bar(fill = "orange", color = "black", alpha = 0.7) +
  labs(
    title = "Previous Loan Defaults Distribution",
    x = "Previous Loan Defaults",
    y = "Frequency"
  ) +
  theme_minimal()

# Bar plot for loan status
plot_status <- ggplot(data_cleaned, aes(x = loan_status)) +
  geom_bar(fill = "purple", color = "black", alpha = 0.7) +
  labs(
    title = "Loan Status Distribution",
    x = "Loan Status",
    y = "Frequency"
  ) +
  theme_minimal()

# Combine all the plots in a grid
grid.arrange(
  plot_gender, plot_home, plot_education,
  plot_intent, plot_defaults, plot_status,
  ncol = 2
)

```

Also we can see some variable distribution.

```{r}
# Distribution of Key Features
# Age Distribution
age_plot <- ggplot(data_cleaned, aes(x = person_age)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  ggtitle("Age Distribution") +
  xlab("Age") +
  ylab("Count")

# Income Distribution
income_plot <- ggplot(data_cleaned, aes(x = person_income)) +
  geom_histogram(binwidth = 10000, fill = "green", color = "black", alpha = 0.7) +
  ggtitle("Income Distribution") +
  xlab("Income") +
  ylab("Count")

# Loan Amount Distribution
loan_amount_plot <- ggplot(data_cleaned, aes(x = loan_amnt)) +
  geom_histogram(binwidth = 5000, fill = "orange", color = "black", alpha = 0.7) +
  ggtitle("Loan Amount Distribution") +
  xlab("Loan Amount") +
  ylab("Count")

# Credit Score Distribution
credit_score_plot <- ggplot(data_cleaned, aes(x = credit_score)) +
  geom_histogram(binwidth = 50, fill = "red", color = "black", alpha = 0.7) +
  ggtitle("Credit Score Distribution") +
  xlab("Credit Score") +
  ylab("Count")

# Display all distribution plots
grid.arrange(age_plot, income_plot, loan_amount_plot, credit_score_plot, ncol = 2)
```

**Creating diferents datasetes**

```{r}
data_cleaned_factor <- read.csv("loan_data_cleaned.csv", header=TRUE, stringsAsFactors = TRUE)
```

Dataset with normalization between 0-1 without categorical factors. The
normalization is created with the function:
$$X{\text{norm}} = \frac{X - X{\min}}{X{\max} - X{\min}}$$

```{r}
# Create normalized dataset
normalize <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}

# Create a copy of the original dataset for normalization
data_cleaned_normalized <- data_cleaned

# Apply normalization in the new dataset
data_cleaned_normalized$person_age <- normalize(data_cleaned$person_age)
data_cleaned_normalized$person_income <- normalize(data_cleaned$person_income)
data_cleaned_normalized$person_emp_exp <- normalize(data_cleaned$person_emp_exp)
data_cleaned_normalized$loan_amnt <- normalize(data_cleaned$loan_amnt)
data_cleaned_normalized$loan_int_rate <- normalize(data_cleaned$loan_int_rate)
data_cleaned_normalized$credit_score <- normalize(data_cleaned$credit_score)
data_cleaned_normalized$cb_person_cred_hist_length <- normalize(data_cleaned$cb_person_cred_hist_length)

# Verify the changes in the new dataset
summary(data_cleaned_normalized)
```

Dataset with categorical factors and normalization between 0-1

```{r}
# Create a copy of the original dataset to apply normalization
data_cleaned_factor_normalized <- data_cleaned_factor

# Apply normalization to the new dataset
data_cleaned_factor_normalized$person_age <- normalize(data_cleaned_factor$person_age)
data_cleaned_factor_normalized$person_income <- normalize(data_cleaned_factor$person_income)
data_cleaned_factor_normalized$person_emp_exp <- normalize(data_cleaned_factor$person_emp_exp)
data_cleaned_factor_normalized$loan_amnt <- normalize(data_cleaned_factor$loan_amnt)
data_cleaned_factor_normalized$loan_int_rate <- normalize(data_cleaned_factor$loan_int_rate)
data_cleaned_factor_normalized$credit_score <- normalize(data_cleaned_factor$credit_score)
data_cleaned_factor_normalized$cb_person_cred_hist_length <- normalize(data_cleaned_factor$cb_person_cred_hist_length)

# Verify the changes in the new dataset
summary(data_cleaned_factor_normalized)
```

Dataset normalization with function **scale()** without categorical
factors. The scale function normalize our data this way:
$$X_{\text{scaled}} = \frac{X - \mu}{\sigma}$$

```{r}
# Create normalized dataset using scale() function
numeric_columns <- data_cleaned %>% select_if(is.numeric) %>% select(-loan_status) %>% names()
data_cleaned_normalized2 <- data_cleaned  # Create a copy for scaled data
data_cleaned_normalized2[, numeric_columns] <- scale(data_cleaned[, numeric_columns])

# Verify the summary of the scaled dataset
summary(data_cleaned_normalized2)
```

Dataset normalization with function **scale()** and categorical factors.

```{r}
# Normalization of numeric columns using scale()
# Ensure that only numeric columns are normalized
numeric_columns <- data_cleaned %>% select_if(is.numeric) %>% select(-loan_status) %>% names()
data_cleaned_factor_normalized2 <- data_cleaned_factor  # Create a copy for scaled data
data_cleaned_factor_normalized2[, numeric_columns] <- scale(data_cleaned[, numeric_columns])

# Verify the summary of the scaled dataset
summary(data_cleaned_factor_normalized2)
```

## **3. Classification methods**

With our pre-processed data, we are now ready to develop and evaluate
classification models to predict loan approval. For this study, we will
use three different machine learning models, each representing distinct
approaches to classification. The chosen models are:

**1. K-Nearest Neighbors (K-NN)**: A distance-based method that
classifies an instance based on its proximity to the k nearest neighbors
in the data-set.

**2. Naïve Bayes**: A probabilistic classifier that assumes independence
between attributes, offering a simple yet effective solution for many
problems.

**3. Decision trees**: A tree-based method that splits data recursively
based on attribute values, providing an interpretable model with clear
decision rules.

To ensure a robust evaluation of these models, we will use two
data-splitting techniques:

**1. Hold out**: A simple method where the data-set is divided into a
single training set and test set, typically using an 80/20 split.

**2. K-fold**: A more comprehensive approach where the data-set is
divided into k subsets (folds), and the model is trained and tested
iteratively on different combinations of these folds.

After training and testing the models using these methods, we will
perform a comparative analysis to assess their performance. Key
evaluation metrics such as accuracy, F1-score, and ROC-AUC will be used
to determine the strengths and weaknesses of each model.

### **3.1. Data Preparation and Normalization**

In this section, we prepare the cleaned dataset for training and testing
by dividing it into appropriate subsets with hold-out method.We will use
two different type of normalization, but it only has repercussion in the
K-NN Model. We have proved that in Tree Decision Model and Naive Bayes,
the type of normalization doesn't have influence in the results.

**First hold-out**

Uses normalized data (with normalize() function) and removes categorical
variables, leaving only numerical predictors.

```{r}
# 1. Separate features (X) and the target variable (y)
X <- data_cleaned_normalized %>% select(-loan_status)  # All columns except the target variable
y <- as.factor(data_cleaned_normalized$loan_status)   # Convert the target variable into a factor (for classification)

# 2. Remove categorical variables from X
X <- X %>% select_if(is.numeric)  # Select only numeric columns

# 3. Split the data into training (80%) and test (20%) sets
set.seed(42)
train_index <- createDataPartition(y, p = 0.8, list = FALSE)
X_train <- X[train_index, ]
X_test <- X[-train_index, ]
y_train <- y[train_index]
y_test <- y[-train_index]

# Verify the dimensions of the resulting datasets
cat("Dimensions of X_train:", dim(X_train), "\n")
cat("Dimensions of X_test:", dim(X_test), "\n")
cat("Dimensions of y_train:", length(y_train), "\n")
cat("Dimensions of y_test:", length(y_test), "\n")
```

**Second hold-out**

Uses normalized data (with normalize() funcion) with categorical
variables converted into factors, keeping them as predictors.

```{r}
# 1. Separate features (X) and the target variable (y)
X <- data_cleaned_factor_normalized %>% select(-loan_status)  # All columns except the target variable
y <- as.factor(data_cleaned_factor_normalized$loan_status)   # Convert the target variable into a factor (for classification)

# 4. Split the data into training (80%) and test (20%) sets
set.seed(42)
train_index <- createDataPartition(y, p = 0.8, list = FALSE)
X_train2 <- X[train_index, ]
X_test2 <- X[-train_index, ]
y_train2 <- y[train_index]
y_test2 <- y[-train_index]

# Verify the dimensions of the resulting datasets
cat("Dimensions of X_train:", dim(X_train2), "\n")
cat("Dimensions of X_test:", dim(X_test2), "\n")
cat("Dimensions of y_train:", length(y_train2), "\n")
cat("Dimensions of y_test:", length(y_test2), "\n")
```

**Third hold-out**

Similar to the first, but with a second set of normalized data (with
scale() function), again removing categorical variables to compare the
effects of different normalizations.

```{r}
# 1. Separate features (X) and the target variable (y)
X <- data_cleaned_normalized2 %>% select(-loan_status)  # All columns except the target variable
y <- as.factor(data_cleaned_normalized2$loan_status)   # Convert the target variable into a factor (for classification)

# 2. Remove categorical variables from X
X <- X %>% select_if(is.numeric)  # Select only numeric columns

# 3. Split the data into training (80%) and test (20%) sets
set.seed(42)
train_index <- createDataPartition(y, p = 0.8, list = FALSE)
X_train3 <- X[train_index, ]
X_test3 <- X[-train_index, ]
y_train3 <- y[train_index]
y_test3 <- y[-train_index]

# Verify the dimensions of the resulting datasets
cat("Dimensions of X_train:", dim(X_train3), "\n")
cat("Dimensions of X_test:", dim(X_test3), "\n")
cat("Dimensions of y_train:", length(y_train3), "\n")
cat("Dimensions of y_test:", length(y_test3), "\n")

```

**Fourth hold-out**

Similar to the second, but based on the second dataset (with scale()
function), keeping categorical variables as factors after the second
normalization.

```{r}
# 1. Separate features (X) and the target variable (y)
X <- data_cleaned_factor_normalized2 %>% select(-loan_status)  # All columns except the target variable
y <- as.factor(data_cleaned_factor_normalized2$loan_status)   # Convert the target variable into a factor (for classification)

# 2. Split the data into training (80%) and test (20%) sets
set.seed(42)
train_index <- createDataPartition(y, p = 0.8, list = FALSE)
X_train4 <- X[train_index, ]
X_test4 <- X[-train_index, ]
y_train4 <- y[train_index]
y_test4 <- y[-train_index]

# Verify the dimensions of the resulting datasets
cat("Dimensions of X_train:", dim(X_train4), "\n")
cat("Dimensions of X_test:", dim(X_test4), "\n")
cat("Dimensions of y_train:", length(y_train4), "\n")
cat("Dimensions of y_test:", length(y_test4), "\n")
```

### **3.2. Training Models**

#### **3.2.1. K-Nearest Neighbors (K-NN)**

[**With K-Fold Cross validation**]{.underline}

In this model with Cross validation we can find out the best **K** for
the model and in the following experiments we will use this one. We do
this because finding the hyper-parameter for each model would be very
computationally expensive.

```{r}
set.seed(42)

# Step 1: Use dummyVars() to convert categorical variables to One-Hot encoding
dummies_model <- dummyVars(" ~ .", data = X_train4, fullRank = TRUE)

# Transform the training data using predict()
X_train4_encoded <- predict(dummies_model, newdata = X_train4)

# Apply the same transformation to the test data
X_test4_encoded <- predict(dummies_model, newdata = X_test4)

# Note: X_train4_encoded and X_test4_encoded are now matrices, you can convert them to data frames if preferred
X_train4_encoded <- as.data.frame(X_train4_encoded)
X_test4_encoded <- as.data.frame(X_test4_encoded)

# Step 2: Set up K-Fold cross-validation
k <- 5
kfold_control <- trainControl(method = "cv", number = k)

# Step 3: Train the K-NN model with K-Fold using the encoded data
knn_kfold_model <- train(
  x = X_train4_encoded,  # Use observations with encoded variables for cross-validation
  y = y_train4,
  method = "knn",
  tuneGrid = expand.grid(k = seq(1, 15, by = 2)),  # Test different values of k
  trControl = kfold_control,
  metric = "Accuracy"
)

# Step 4: Print details of the trained model
print(knn_kfold_model)
cat("Best k value (K-Fold):", knn_kfold_model$bestTune$k, "\n")

# Step 5: Make predictions on the test set
# Here we correct to use X_test4_encoded
knn_predictions <- predict(knn_kfold_model, X_test4_encoded)

# Step 6: Evaluate the model on the test set
confusion <- confusionMatrix(knn_predictions, y_test4)
print(confusion)

```

[**Without K-Fold Cross validation**]{.underline}

**Without categorical factors** and using **normalization between 0-1**.
Also we assume k = 13.

```{r}
set.seed(42)

# Select a fixed value for k (adjust as needed)
best_k <- 13  # Assuming this is the best k value found previously

# Train the K-NN model with a single k value
knn_model <- train(
  x = X_train,
  y = y_train,
  method = "knn",
  tuneGrid = expand.grid(k = best_k),  # Single value for k
  metric = "Accuracy",
  trControl = trainControl(method = "none")  # No cross-validation
)

# Print details of the trained model
print(knn_model)

# Make predictions on the test set
knn_predictions <- predict(knn_model, X_test)

# Evaluate the model on the test set
confusion <- confusionMatrix(knn_predictions, y_test)
print(confusion)
```

**With categorical transform and normalization (0-1):**

**One-hot encoding** is a technique used to convert categorical
variables into a format that machine learning models can use. Models
cannot directly handle text labels or categories because they require
numerical values as input. After we transform the categorical data we do
**one-hot encoding**, to use this transform data in the K-NN model.

```{r}
set.seed(42)

# Assume your training and testing datasets are as follows:
# X_train2 contains the training predictors
# y_train2 contains the training target variable

# Step 1: Use dummyVars() to convert categorical variables into One-Hot encoding
dummies_model <- dummyVars(" ~ .", data = X_train2, fullRank = TRUE)

# Transform the training data using predict()
X_train2_encoded <- predict(dummies_model, newdata = X_train2)

# Apply the same transformation to the test data
X_test2_encoded <- predict(dummies_model, newdata = X_test2)

# Note: X_train2_encoded and X_test2_encoded are now matrices; you can convert them to data frames if preferred
X_train2_encoded <- as.data.frame(X_train2_encoded)
X_test2_encoded <- as.data.frame(X_test2_encoded)

# Step 2: Train the K-NN model with the selected k value

# Train the K-NN model with a single k value
knn_model <- train(
  x = X_train2_encoded,
  y = y_train2,
  method = "knn",
  tuneGrid = expand.grid(k = best_k),  # Single value for k
  metric = "Accuracy",
  trControl = trainControl(method = "none")  # No cross-validation
)

# Print details of the trained model
print(knn_model)

# Make predictions on the test set
knn_predictions <- predict(knn_model, X_test2_encoded)

# Evaluate the model on the test set
confusion <- confusionMatrix(knn_predictions, y_test2)
print(confusion)
```

**Without categorical data** and using **normalization with function**
**scale()**:

```{r}
set.seed(42)

# Select a fixed value for k (adjust as needed)
best_k <- 13  # Assuming this is the best k value found previously

# Train the K-NN model with a single k value
knn_model <- train(
  x = X_train3,
  y = y_train3,
  method = "knn",
  tuneGrid = expand.grid(k = best_k),  # Single value for k
  metric = "Accuracy",
  trControl = trainControl(method = "none")  # No cross-validation
)

# Print details of the trained model
print(knn_model)

# Make predictions on the test set
knn_predictions <- predict(knn_model, X_test3)

# Evaluate the model on the test set
confusion <- confusionMatrix(knn_predictions, y_test3)
print(confusion)
```

**With categorical transform** and **normalization with function
scale()**:

Also we do **one-hot encoding**

```{r}
set.seed(42)

# Assume your training and testing datasets are:
# X_train4 contains the training predictors
# y_train4 contains the training target variable

# Step 1: Use dummyVars() to convert categorical variables into One-Hot encoding
dummies_model <- dummyVars(" ~ .", data = X_train4, fullRank = TRUE)

# Transform the training data using predict()
X_train4_encoded <- predict(dummies_model, newdata = X_train4)

# Apply the same transformation to the test data
X_test4_encoded <- predict(dummies_model, newdata = X_test4)

# Note: X_train4_encoded and X_test4_encoded are now matrices; you can convert them to data frames if preferred
X_train4_encoded <- as.data.frame(X_train4_encoded)
X_test4_encoded <- as.data.frame(X_test4_encoded)

# Step 2: Train the K-NN model with the selected k value

# Train the K-NN model with a single k value
knn_model <- train(
  x = X_train4_encoded,
  y = y_train4,
  method = "knn",
  tuneGrid = expand.grid(k = best_k),  # Single value for k
  metric = "Accuracy",
  trControl = trainControl(method = "none")  # No cross-validation
)

# Print details of the trained model
print(knn_model)

# Make predictions on the test set
knn_predictions <- predict(knn_model, X_test4_encoded)

# Evaluate the model on the test set
confusion <- confusionMatrix(knn_predictions, y_test4)
print(confusion)
```

**Conclusion**

From the results of the five k-NN models, the following observations can
be made:

1.  The model with k-fold cross-validation performed the best overall,
    achieving the highest accuracy (89.65%) and balanced accuracy
    (82.05%) when using 22 predictors and the optimal k=15.
    Cross-validation likely improved generalizability by mitigating
    over-fitting.

2.  Models without categorical data generally outperformed those with
    categorical transformations, as the categorical versions had lower
    accuracy and kappa values, showing limited added value from
    categorical variables.

3.  Using normalization (0-1) consistently yielded higher accuracy and
    kappa scores compared to the scale function, indicating that feature
    scaling methods can influence model performance significantly.

4.  Models with fewer predictors (8) performed worse, showing reduced
    accuracy (\~83.4%) and balanced accuracy (\~70.5%), likely due to
    loss of information after excluding many features.

5.  Overall, normalization and including all numerical predictors (22
    features) yielded the most effective results, especially when
    combined with cross-validation, underscoring the importance of
    feature selection and validation strategies in k-NN performance.

#### **3.2.2. Naïve Bayes**

[**Without K-Fold Cross Validation**]{.underline}

Just as it has been done in the case of K-NN, training 4 models with
different datasets, we now apply it with Naïve Bayes.

**Without categorical data** and using **normalization between 0-1**:

```{r}
# Train the Naive Bayes model with Laplace smoothing
set.seed(42)
naive_model <- naiveBayes(x = X_train, y = y_train, laplace = 1)

# Make predictions on the test set
naive_predictions <- predict(naive_model, X_test)

# Confusion matrix to evaluate the model
confusion <- confusionMatrix(naive_predictions, y_test)
print(confusion)
```

**With categorical transform and normalization (0-1):**

```{r}
# Train the Naive Bayes model with Laplace smoothing
set.seed(42)
naive_model <- naiveBayes(x = X_train2, y = y_train2, laplace = 1)

# Make predictions on the test set
naive_predictions <- predict(naive_model, X_test2)

# Confusion matrix to evaluate the model
confusion <- confusionMatrix(naive_predictions, y_test2)
print(confusion)
```

**Without categorical data** and using **normalization with function
scale()**:

```{r}
# Train the Naive Bayes model with Laplace smoothing
set.seed(42)
naive_model <- naiveBayes(x = X_train3, y = y_train3, laplace = 1)

# Make predictions on the test set
naive_predictions <- predict(naive_model, X_test3)

# Confusion matrix to evaluate the model
confusion <- confusionMatrix(naive_predictions, y_test3)
print(confusion)
```

**With categorical transform** and **normalization with function
scale()**:

```{r}
# Train the Naive Bayes model with Laplace smoothing
set.seed(42)
naive_model <- naiveBayes(x = X_train4, y = y_train4, laplace = 1)

# Make predictions on the test set
naive_predictions <- predict(naive_model, X_test4)

# Confusion matrix to evaluate the model
confusion <- confusionMatrix(naive_predictions, y_test4)
print(confusion)
```

[**With K-Fold Cross Validation**]{.underline}

In this case, we aimed to use the K-Fold Cross-Validation method, but
encountered an issue with several warnings such as "Numerical 0
probability for all classes with observation." After investigating the
issue, we concluded that it is likely due to the imbalance in the target
variable, where class 0 has significantly more observations than
class 1. As a result, when the model creates the folds, some folds may
contain very few or no examples of class 1, leading to this warning.
(Finally we removed the Warnings in the chunk, and see anyways how
affect to the model)

```{r warning=FALSE}
# Configure K-Fold cross-validation
set.seed(42)

# Use caret with K-Fold and SMOTE
kfold_control <- trainControl(method = "cv", number = 5)

# Train the Naive Bayes model using K-Fold cross-validation
naive_kfold_model <- train(
  x = X_train,
  y = y_train,
  method = "nb",
  trControl = kfold_control,
  metric = "Accuracy",
  tuneGrid = expand.grid(fL = 1, usekernel = FALSE, adjust = 1)  # With smoothing
)

# Print details of the trained model
print(naive_kfold_model)

# Make predictions on the test set
naive_kfold_predictions <- predict(naive_kfold_model, X_test)

# Confusion matrix to evaluate the model on the test set
confusion <- confusionMatrix(naive_kfold_predictions, y_test)
print(confusion)
```

**Conclusion**

The Naive Bayes models exhibit varied performance based on the
preprocessing and validation approach used:

1.  Models without categorical data and normalization (0-1) show a lower
    accuracy (\~81.92%) and Kappa (\~0.43), indicating moderate
    agreement and predictive ability. The Balanced Accuracy (\~69.57%)
    highlights suboptimal specificity (47.35%).

2.  Models with categorical transformation and normalization (0-1)
    achieve a higher accuracy (\~88.22%) and Kappa (\~0.66). This
    improvement stems from better integration of categorical variables,
    resulting in a Balanced Accuracy of 82.82% and enhanced specificity
    (73.10%).

3.  Cross-validated Naive Bayes (k-fold) achieves similar performance to
    the single hold-out models without categorical data, maintaining
    accuracy (81.92%) and consistent sensitivity (\~91.8%) but low
    specificity (47.35%).

These results demonstrate that incorporating categorical data
significantly improves Naive Bayes performance, with normalization
contributing modestly. K-fold validation ensures generalizability but
does not inherently enhance the model’s accuracy compared to the best
single hold-out scenarios.

#### **3.2.3. Decision trees**

[**Without K-Fold Cross Validation**]{.underline}

Just as it was done in the case of K-NN,and Naïve Bayes where 4 models
were trained using different data-sets, we now apply the same approach
with decision trees. Four separate models are trained and evaluated,
considering variations in data-sets: with and without categorical data,
as well as with and without normalization.

**Without categorical data**

```{r}
# Train the decision tree model
set.seed(42)
decision_tree_model <- rpart(
  formula = y_train ~ .,  # Formula: y_train as the dependent variable and all columns of X_train as independent variables
  data = data.frame(X_train, y_train),  # Create a data frame combining X_train and y_train
  method = "class"  # Method for classification
)

# Model summary
print(decision_tree_model)

# Visualize the decision tree
rpart.plot(decision_tree_model, type = 2, extra = 102, main = "Decision Tree")

# Predictions on the test set
decision_tree_predictions <- predict(decision_tree_model, data.frame(X_test), type = "class")

# Model evaluation
confusion <- confusionMatrix(decision_tree_predictions, y_test)
print(confusion)

```

**With categorical data**

```{r}
library(rpart)
library(rpart.plot)

# Train the decision tree model
set.seed(42)
decision_tree_model <- rpart(
  formula = y_train2 ~ .,  # Formula: y_train2 as the dependent variable and all columns of X_train2 as independent variables
  data = data.frame(X_train2, y_train2),  # Create a data frame combining X_train2 and y_train2
  method = "class"  # Method for classification
)

# Model summary
print(decision_tree_model)

# Visualize the decision tree
rpart.plot(decision_tree_model, type = 2, extra = 102, main = "Decision Tree")

# Predictions on the test set
decision_tree_predictions <- predict(decision_tree_model, data.frame(X_test2), type = "class")

# Model evaluation
confusion <- confusionMatrix(decision_tree_predictions, y_test2)
print(confusion)

```

[**With K-Fold Cross Validation**]{.underline}

In this case, we conducted two experiments: one without categorical data
and another including categorical data.

**Without categorical data**

```{r}
# Set up K-Fold cross-validation
set.seed(42)
kfold_control <- trainControl(method = "cv", number = 5)

# Train the decision tree model with K-Fold
set.seed(42)
decision_tree_model <- train(
  x = X_train,
  y = y_train,
  method = "rpart",  # Method for decision tree in caret
  trControl = kfold_control,
  metric = "Accuracy",  # Evaluation metric
  tuneGrid = expand.grid(cp = seq(0.01, 0.1, by = 0.01))  # Test different complexity parameter (cp) values
)

# Print the model results
print(decision_tree_model)

# Best hyperparameter (cp)
cat("Best cp value:", decision_tree_model$bestTune$cp, "\n")

# Predictions on the test set
tree_predictions <- predict(decision_tree_model, X_test)

# Model evaluation
confusion <- confusionMatrix(tree_predictions, y_test)
print(confusion)

```

**With categorical data**

```{r}
# Set up K-Fold cross-validation
set.seed(42)
kfold_control <- trainControl(method = "cv", number = 5)

# Train the decision tree model with K-Fold
set.seed(42)
decision_tree_model <- train(
  x = X_train2,
  y = y_train2,
  method = "rpart",  # Method for decision tree in caret
  trControl = kfold_control,
  metric = "Accuracy",  # Evaluation metric
  tuneGrid = expand.grid(cp = seq(0.01, 0.1, by = 0.01))  # Test different complexity parameter (cp) values
)

# Print the model results
print(decision_tree_model)

# Best hyperparameter (cp)
cat("Best cp value:", decision_tree_model$bestTune$cp, "\n")

# Predictions on the test set
tree_predictions <- predict(decision_tree_model, X_test2)

# Model evaluation
confusion <- confusionMatrix(tree_predictions, y_test2)
print(confusion)

```

**Conclusion**

The results show that decision tree models improve significantly when
categorical data is included. In models without categorical variables,
sensitivity is high (92.77%), but specificity is low (58.00%),
indicating an imbalance in classification. When categorical variables
are incorporated, both sensitivity (96.90%) and specificity (71.55%)
increase significantly, resulting in a more balanced accuracy (91.26%
accuracy, Balanced Accuracy = 84.22%). This highlights the importance of
considering all relevant variables in the model. Additionally, higher
Kappa values with categorical data reflect better consistency in
predictions. Finally, cross-validation confirms the stability of the
model's performance.

## **4. Main Conclusion**

This assignment set out to develop predictive models to classify loan
approval statuses by thoroughly analyzing and processing a dataset,
followed by implementing machine learning techniques. Through the
process, a comprehensive understanding of the dataset and the models was
achieved, leading to insightful conclusions and the successful
completion of the proposed objectives.

The first step involved a detailed Exploratory Data Analysis (EDA). Key
features were analyzed for distributions, correlations, and potential
anomalies. Missing values were addressed, and outliers—such as
unrealistic ages and excessive work experience—were identified and
removed. Relationships between variables were explored, revealing
significant correlations, like the one between age and credit history
length, and their impact on the model's performance.

Data preprocessing played a crucial role in ensuring model readiness.
The dataset was prepared in various forms: normalized, standardized, and
with categorical variables either encoded or left as factors. This
enabled the comparison of how different data transformations affected
model performance. Additionally, redundant variables and correlated
features were handled strategically to minimize biases and
multicollinearity.

With clean and prepared datasets, three machine learning models were
implemented and evaluated:

1.  **K-Nearest Neighbors (K-NN)**: This model benefited significantly
    from normalization and hyperparameter tuning through k-fold
    cross-validation, achieving a high accuracy of 89.65%. However,
    incorporating categorical variables did not yield substantial
    improvements, highlighting the model's reliance on numerical
    features.

2.  **Naïve Bayes**: This model demonstrated notable improvements when
    categorical data was included, achieving an accuracy of 88.22%. Its
    reliance on feature diversity emphasized the importance of robust
    feature engineering.

3.  **Decision Trees**: This model outperformed others, achieving an
    accuracy of 91.26% with categorical data. The tree's balance of
    sensitivity and specificity made it a standout choice, particularly
    when combined with k-fold cross-validation for fine-tuning.

The evaluation of these models revealed that **decision trees** were the
most effective in balancing predictive performance and interpretability,
making them ideal for this classification task. K-fold cross-validation
emerged as an invaluable technique for robust model assessment and
hyperparameter optimization.

The assignment provided critical insights into the importance of data
preparation and feature engineering. Addressing outliers and leveraging
the full potential of categorical variables were decisive factors in the
success of the models. Moreover, the comparison between normalization
techniques underscored the impact of preprocessing on models like K-NN,
which rely on distance metrics.

In conclusion, this project achieved its objectives by constructing and
evaluating predictive models for loan approval classification. The
decision tree model, particularly when trained with categorical
variables, proved to be the best-performing and most interpretable
solution. This process reinforced the value of systematic data analysis,
preprocessing, and validation in developing effective machine learning
models.
