---
title: "Final practice"
author: Antonio Torres Escribano, Jaume Cantallops Comas, Jordi Florit Ensenyat, Josep
  Gabriel Fornes Reynes and Pau Girón Rodríguez
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

## 1. Introduction

Real estate investment is a cornerstone of financial growth, yet it
remains a complex and often unpredictable field. Investors frequently
struggle to identify profitable opportunities due to the overwhelming
number of factors influencing property values, as well as the lack of
systematic, data-driven approaches to decision-making.

The primary objective of this research is to determine the key
characteristics that should guide investment decisions in residential
properties across Portugal. By analyzing patterns within real estate
data, this study aims to empower investors to make more informed and
strategic choices, reducing the risks associated with traditional
methods based on intuition or incomplete information.

To tackle this challenge, the problem has been divided into three
specific sub-questions, each contributing to the overall understanding
of the problem:

1.  Identifying the factors that most influence property prices: What
    variables, such as location, play the largest role in determining a
    property’s value?

2.  Recognizing high-value property traits: What common characteristics
    are shared by properties with higher prices, and how can these
    traits be used to identify similar investment opportunities?

3.  Highlighting areas with undervalued properties: Which regions in
    Portugal have a higher density of properties that are priced below
    their true market value, offering potential for high returns?

By addressing these questions, this study not only seeks to provide
clarity on the drivers of property value but also aims to uncover hidden
opportunities in the market. The use of data-driven methodologies
ensures that the analysis is objective, reliable, and capable of
adapting to the dynamic nature of real estate trends.

Ultimately, this research will serve as a practical guide for investors
looking to maximize their returns in the Portuguese housing market,
demonstrating the transformative potential of data science in solving
real-world problems.

# 2. Exploratory Data Analysis (EDA)

## 2.1 Loading and Inspecting the Dataset

```{r message=FALSE}
library(visdat)
library(tidyr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(ggcorrplot)
library(corrplot)
library(cluster)
```

```{r}
portugal_data <- read.csv(file = 'portugal_housing.csv')
```

```{r}
str(portugal_data)
```

## 2.2 Missing values

Here we will see how there are columns that, instead of having NA,
contain empty values ("").

```{r}
head(levels(as.factor(portugal_data$Town)))
```

We replace the variables that contain `Ã\u0081zere` with Azere.

```{r}
portugal_data <- portugal_data %>%
  mutate(Town = if_else(Town == "Ã\u0081zere", "Azere", Town))
```

Then we'll check the categorical values as factors

```{r}
levels(as.factor(portugal_data$Type))
```

```{r}
levels(as.factor(portugal_data$EnergyCertificate))
```

```{r}
levels(as.factor(portugal_data$Floor))
```

```{r}
levels(as.factor(portugal_data$Lift))
```

```{r}
levels(as.factor(portugal_data$HasParking))
```

```{r}
head(levels(as.factor(portugal_data$PublishDate)))
```

```{r}
levels(as.factor(portugal_data$Garage))
```

```{r}
levels(as.factor(portugal_data$Elevator))
```

```{r}
levels(as.factor(portugal_data$ElectricCarsCharging))
```

```{r}
levels(as.factor(portugal_data$ConservationStatus))
```

Instead of having empty values, we will replace these values with NA:

```{r}
portugal_data$Town[portugal_data$Town == ""] <- NA
summary(as.numeric(as.factor(portugal_data$Town)))
```

```{r}
portugal_data$Type[portugal_data$Type == ""] <- NA
summary(as.numeric(as.factor(portugal_data$Type)))
```

For EnergyCertificate instead of assigning NA to the empty values, we
assign 'No Certificate', same for the values 'Not available' and 'NC'.

```{r}
portugal_data$EnergyCertificate[portugal_data$EnergyCertificate == ""] <- "No Certificate"
portugal_data$EnergyCertificate[portugal_data$EnergyCertificate == "Not available"] <- "No Certificate"
portugal_data$EnergyCertificate[portugal_data$EnergyCertificate == "NC"] <- "No Certificate"
summary(as.numeric(as.factor(portugal_data$EnergyCertificate)))
```

```{r}
portugal_data$Floor[portugal_data$Floor == ""] <- NA
summary(as.numeric(as.factor(portugal_data$Floor)))
```

```{r}
portugal_data$Lift[portugal_data$Lift == ""] <- NA
summary(as.numeric(as.factor(portugal_data$Lift)))
```

```{r}
portugal_data$HasParking[portugal_data$HasParking == ""] <- NA
summary(as.numeric(as.factor(portugal_data$HasParking)))
```

```{r}
portugal_data$PublishDate[portugal_data$PublishDate == ""] <- NA
summary(as.numeric(as.factor(portugal_data$PublishDate)))
```

```{r}
portugal_data$Garage[portugal_data$Garage == ""] <- NA
summary(as.numeric(as.factor(portugal_data$Garage)))
```

```{r}
portugal_data$Elevator[portugal_data$Elevator == ""] <- NA
summary(as.numeric(as.factor(portugal_data$Elevator)))
```

```{r}
portugal_data$ElectricCarsCharging[portugal_data$ElectricCarsCharging == ""] <- NA
summary(as.numeric(as.factor(portugal_data$ElectricCarsCharging)))
```

```{r}
portugal_data$ConservationStatus[portugal_data$ConservationStatus == ""] <- NA
summary(as.numeric(as.factor(portugal_data$ConservationStatus)))
```

We'll convert all categorical columns to numeric in the dataset.

```{r}
portugal_data$District <- as.factor(portugal_data$District)
portugal_data$City <- as.factor(portugal_data$City)
portugal_data$Town <- as.factor(portugal_data$Town)
portugal_data$Type <- as.factor(portugal_data$Type)
portugal_data$EnergyCertificate <- as.factor(portugal_data$EnergyCertificate)
portugal_data$Floor <- as.factor(portugal_data$Floor)
portugal_data$Lift <- as.factor(portugal_data$Lift)
portugal_data$HasParking <- as.factor(portugal_data$HasParking)
portugal_data$PublishDate <- as.factor(portugal_data$PublishDate)
portugal_data$Garage <- as.factor(portugal_data$Garage)
portugal_data$Elevator <- as.factor(portugal_data$Elevator)
portugal_data$ElectricCarsCharging <- as.factor(portugal_data$ElectricCarsCharging)
portugal_data$ConservationStatus <- as.factor(portugal_data$ConservationStatus)

summary(portugal_data)
```

Graph of amount of property types

```{r}
ggplot(portugal_data, aes(x = Type)) +
  geom_bar(fill = "#69b3a2") +
  labs(title = "Number of Properties by Type", x = "Property Type", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

We mainly have apartment, house and land.

Next we look at missing values:

```{r}
sapply(portugal_data, function(x) sum(is.na(x)))
sum(is.na(portugal_data))  
```

Percentage of missing values per column

```{r}
sapply(portugal_data, function(x) mean(is.na(x)) * 100)
```

```{r}
missing_percentage <- sapply(portugal_data, function(x) mean(is.na(x)) * 100)

missing_df <- data.frame(
  Variable = names(missing_percentage),
  MissingPercentage = missing_percentage
)

# Plot percentage of missing values
ggplot(missing_df, aes(x = reorder(Variable, -MissingPercentage), y = MissingPercentage)) +
  geom_bar(stat = "identity", fill = "#f46a6a") +
  labs(title = "Percentage of Missing Values by Column", x = "Variable", y = "Percentage (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

If we were to remove rows with missing data, we'd end with no rows at
all.

```{r}
portugal_data_complete_cases <- na.omit(portugal_data)
sum(is.na(portugal_data_complete_cases))  # See the missing value now
dim(portugal_data_complete_cases)    # Get the dimensions of the complete dataset (rows and columns)
```

We visualize the missing values per variable.

```{r}
visdat::vis_dat(portugal_data, warn_large_data = FALSE, large_data_size = 115000)
```

We detect that there are two columns that mean the same thing (Lift and
Elevator). It seems that the rows where one column has NA, the other
contains values, and vice versa. Therefore, we check which rows have
matching NA values and which rows contain values in both columns.

```{r}
# Filter the ones that have NA in both rows
na_coincidencia <- portugal_data[is.na(portugal_data$Lift) & is.na(portugal_data$Elevator), ]

# Show the rows with NA in both columns
nrow(na_coincidencia)
```

```{r}
# Filter rows where both columns have NA
na_coincidencia <- portugal_data[!is.na(portugal_data$Lift) & !is.na(portugal_data$Elevator), ]

# Show the rows with NA in both columns
nrow(na_coincidencia)
```

Therefore, we can see that there are 32 rows where both columns contain
NA values and 0 rows where both contain values. We can conclude that the
two columns have the same meaning and represent the same thing.
Therefore, we can combine them into a single column. When we detect
values in one column, we keep that value, but when it is an NA value, we
will take the value from the other column.

```{r}
# Create a new combined column where we prioritize the NA value
portugal_data <- portugal_data %>%
  mutate(CombinedElevator = case_when(
    !is.na(Lift) ~ as.character(Lift),
    TRUE ~ as.character(Elevator)  # Prioritize Elevator if Lift is NA
  ))

# Convert CombinedElevator to a factor
portugal_data$CombinedElevator <- as.factor(portugal_data$CombinedElevator)

# Remove the original columns if they are no longer needed
portugal_data <- portugal_data %>% select(-Lift, -Elevator)

# Check the result
head(portugal_data)
```

We can see a strong correlation between Parking and HasParking. For this
reason, we will remove the HasParking column. Additionally, it is worth
noting that HasParking is a boolean variable indicating whether parking
is available or not, while Parking is a numeric value indicating the
number of parking spaces available. Therefore, the HasParking column is
unnecessary, as the presence of parking can already be determined by the
number of spaces.

```{r}
# Deleted not needed HasParking column
portugal_data$HasParking <- NULL
summary(portugal_data)
```

We'll separate into the 3 biggest type of real estate, it will help us
treating missing values and outliers, since values can differ a lot
between a House and a Land.

```{r}
dataset_apartment <- subset(portugal_data, Type == "Apartment")
dataset_land <- subset(portugal_data, Type == "Land")
dataset_house <- subset(portugal_data, Type == "House")
```

We visualize the missing values by percentage for each of the types.

```{r}
missing_percentage <- sapply(dataset_apartment, function(x) mean(is.na(x)) * 100)

missing_df <- data.frame(
  Variable = names(missing_percentage),
  MissingPercentage = missing_percentage
)

# Plot percentage of missing values
ggplot(missing_df, aes(x = reorder(Variable, -MissingPercentage), y = MissingPercentage)) +
  geom_bar(stat = "identity", fill = "#f46a6a") +
  labs(title = "Percentage of Missing Values by Column Apartment", x = "Variable", y = "Percentage (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
missing_percentage <- sapply(dataset_apartment, function(x) mean(is.na(x)) * 100)

missing_df <- data.frame(
  Variable = names(missing_percentage),
  MissingPercentage = missing_percentage
)

# Plot percentage of missing values
ggplot(missing_df, aes(x = reorder(Variable, -MissingPercentage), y = MissingPercentage)) +
  geom_bar(stat = "identity", fill = "#f46a6a") +
  labs(title = "Percentage of Missing Values by Column Land", x = "Variable", y = "Percentage (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
missing_percentage <- sapply(dataset_house, function(x) mean(is.na(x)) * 100)

missing_df <- data.frame(
  Variable = names(missing_percentage),
  MissingPercentage = missing_percentage
)

# Plot percentage of missing values
ggplot(missing_df, aes(x = reorder(Variable, -MissingPercentage), y = MissingPercentage)) +
  geom_bar(stat = "identity", fill = "#f46a6a") +
  labs(title = "Percentage of Missing Values by Column Apartment", x = "Variable", y = "Percentage (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

We can see how the most common columns with missing values differ
between types, meaning it's a good decision to split the dataset in 3
different ones.

For better visualization we'll show these percentages for all 3 types in
the same table.

```{r}
# Calculate missing percentages for all three datasets
missing_percentages <- data.frame(
  Variable = names(dataset_apartment),
  Apartment = sapply(dataset_apartment, function(x) mean(is.na(x)) * 100),
  Land = sapply(dataset_land, function(x) mean(is.na(x)) * 100),
  House = sapply(dataset_house, function(x) mean(is.na(x)) * 100)
)

# Print the summary table
print(missing_percentages)

# Highlight columns that are completely empty in one dataset but not others
missing_percentages$CompletelyEmpty <- with(missing_percentages, 
  ifelse(Apartment == 100 & Land < 100 & House < 100, "Apartment", 
  ifelse(Land == 100 & Apartment < 100 & House < 100, "Land",
  ifelse(House == 100 & Apartment < 100 & Land < 100, "House", NA)))
)

print(missing_percentages)
```

We can see that there are variables in certain datasets that no longer
have as many NAs. For example, Floor for apartments, ConstructionYear
for both apartments and houses, among others. However, LivingArea still
has 71% NAs in the land dataset.

Next we'll see a graph with the missing value columns of all 3 types in
the same graph.

```{r}
# Calculate missing percentages for all three datasets
missing_percentages <- data.frame(
  Variable = names(dataset_apartment),
  Apartment = sapply(dataset_apartment, function(x) mean(is.na(x)) * 100),
  Land = sapply(dataset_land, function(x) mean(is.na(x)) * 100),
  House = sapply(dataset_house, function(x) mean(is.na(x)) * 100)
)

# Identify columns with more than 50% missing values in any dataset
missing_percentages$ToBeDropped <- with(missing_percentages, 
  ifelse(Apartment > 50 | Land > 50 | House > 50, "Yes", "No")
)

# Print the summary table with missing percentages and columns to be dropped
print(missing_percentages)

# Filter only the columns that will be dropped
columns_to_drop <- missing_percentages[missing_percentages$ToBeDropped == "Yes", ]
print("Columns with more than 50% missing values (to be dropped):")
print(columns_to_drop)

# Optionally, visualize the columns with more than 50% missing values
missing_percentages_long <- tidyr::pivot_longer(
  missing_percentages,
  cols = c(Apartment, Land, House),
  names_to = "Dataset",
  values_to = "MissingPercentage"
)

# Plot the missing percentages for all datasets
ggplot(missing_percentages_long, aes(x = Variable, y = MissingPercentage, fill = Dataset)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_hline(yintercept = 50, linetype = "dashed", color = "red", size = 1) +
  labs(
    title = "Missing Percentage for Columns Across Types",
    x = "Variable",
    y = "Missing Percentage (%)",
    fill = "Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Another view on the missing values.

```{r}
# Reshape the data for plotting
missing_melt <- melt(missing_percentages, id.vars = "Variable", 
                     measure.vars = c("Apartment", "Land", "House"),
                     variable.name = "Dataset", value.name = "MissingPercentage")

# Plot the heatmap
ggplot(missing_melt, aes(x = Dataset, y = Variable, fill = MissingPercentage)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "#f46a6a", na.value = "white") +
  labs(title = "Missing Value Comparison Across Datasets", x = "Dataset", y = "Variable") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

We'll remove the columns with a higher percentage than 50 in the
clean_dataset method.

```{r}
handle_outliers_by_type <- function(df, columns) {
  # Remove rows with outliers for the given columns
  df <- df %>%
    filter(
      # Retain rows within the bounds for all specified columns
      rowSums(sapply(df[columns], function(.x) {
        Q1 <- quantile(.x, 0.25, na.rm = TRUE)
        Q3 <- quantile(.x, 0.75, na.rm = TRUE)
        IQR <- Q3 - Q1
        lower_bound <- pmax(0, Q1 - 1.5 * IQR)
        upper_bound <- Q3 + 1.5 * IQR
        # Return TRUE if the value is within the bounds
        !(.x < lower_bound | .x > upper_bound)
      })) == length(columns)
    )

  return(df)
}

clean_dataset <- function(data) {
  # Handle missing values
  missing_percentage <- colSums(is.na(data)) / nrow(data) * 100
  columns_to_drop <- names(missing_percentage[missing_percentage > 50])
  data <- data[, setdiff(names(data), columns_to_drop)]

  # Impute missing numeric values with median
  numeric_cols <- sapply(data, is.numeric)
  data[, numeric_cols] <- lapply(data[, numeric_cols], function(x) {
    ifelse(is.na(x), median(x, na.rm = TRUE), x)
  })
  
  # Impute missing categorical values with the most frequent value
  categorical_cols <- sapply(data, is.factor)
  data[, categorical_cols] <- lapply(data[, categorical_cols], function(x) {
    x[is.na(x)] <- names(sort(table(x), decreasing = TRUE))[1]
    return(as.factor(x))
  })
  
  # Handle outliers for numeric columns
  columns_to_fix <- names(data)[sapply(data, is.numeric)]
  data <- handle_outliers_by_type(data, columns_to_fix)
  
  return(data)
}
```

```{r}
dataset_apartment_test <- clean_dataset(dataset_apartment)

summary(dataset_apartment_test)
```

It didn't remove all the outliers, since we have columns like TotalArea
and LivingArea with a min of 0, which doesn't realistically make sense.
And very low price values, because they have high values, the mean is
very high, which causes the lower bound of 1.5\*IQR to be negative.

```{r}
# Create the histograms
p1 <- ggplot(dataset_land, aes(x = Price)) +
  geom_histogram(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Price Distribution", x = "Price", y = "Frequency") +
  theme_minimal()

p2 <- ggplot(dataset_land, aes(x = log(Price))) +
  geom_histogram(binwidth = 0.1, fill = "orange", color = "black", alpha = 0.7) +
  labs(title = "LogPrice Distribution", x = "Log(Price)", y = "Frequency") +
  theme_minimal()

# Use gridExtra to display them side by side
library(gridExtra)
grid.arrange(p1, p2, ncol = 2)
```

After applying the transformation we can see how it's normalized now.
We'll check the distribution of all other variables.

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

plot_histograms_for_all_columns <- function(data) {
  # Select only numeric columns
  numeric_data <- data %>%
    select(where(is.numeric))
  
  # Convert dataset to long format using pivot_longer
  data_long <- numeric_data %>%
    pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value")
  
  # Plot histograms for all numeric columns
  ggplot(data_long, aes(x = Value)) +
    geom_histogram(bins = 30, fill = "skyblue", color = "black", alpha = 0.7) +
    facet_wrap(~ Variable, scales = "free_x", ncol = 4) +
    theme_minimal() +
    labs(
      title = "Histograms of All Numeric Columns in the Dataset",
      x = "Value",
      y = "Frequency"
    )
}

# Example: Plot histograms for the dataset_land
plot_histograms_for_all_columns(dataset_land)
```

Not a single normalized one, so we'll modify the method to transform the
variables to log before aplying IQR.

```{r}
handle_outliers_with_log <- function(df, columns) {
  # Detect and handle outliers after transformation
  df <- df %>%
    filter(
      rowSums(sapply(df[columns], function(.x) {
        Q1 <- quantile(.x, 0.25, na.rm = TRUE)
        Q3 <- quantile(.x, 0.75, na.rm = TRUE)
        IQR <- Q3 - Q1
        lower_bound <- Q1 - 1.5 * IQR
        upper_bound <- Q3 + 1.5 * IQR
        !(.x < lower_bound | .x > upper_bound)
      })) == length(columns)
    )
  return(df)
}

clean_dataset_with_log <- function(data) {
  # Handle missing values: Drop columns with >50% missing data
  missing_percentage <- colSums(is.na(data)) / nrow(data) * 100
  columns_to_drop <- names(missing_percentage[missing_percentage > 50])
  data <- data[, setdiff(names(data), columns_to_drop)]

  # Identify numeric and categorical columns
  numeric_columns <- names(data)[sapply(data, is.numeric)]
  categorical_columns <- names(data)[sapply(data, is.factor)]

  # Impute missing numeric values with median
  data[, numeric_columns] <- lapply(data[, numeric_columns], function(x) {
    ifelse(is.na(x), median(x, na.rm = TRUE), x)
  })

  # Impute missing categorical values with the most frequent value
  data[, categorical_columns] <- lapply(data[, categorical_columns], function(x) {
    x[is.na(x)] <- names(sort(table(x), decreasing = TRUE))[1]
    return(as.factor(x))
  })

  # Drop rows with negative values in numeric columns
  data <- data %>%
    filter(
      # Keep rows where all numeric columns have non-negative values
      !rowSums(data[, numeric_columns] < 0, na.rm = TRUE) > 0
    )

  # Add log-transformed column for Price
  data <- data %>%
    mutate(across(all_of(numeric_columns), ~ log(.x + 1), .names = "Log_{.col}"))

  numeric_columns <- names(data)[sapply(data, is.numeric)]

  # Handle outliers for numeric columns (including LogPriceXd)
  data <- handle_outliers_with_log(data, numeric_columns)

  log_columns <- paste("Log", numeric_columns, sep = "_")  # Create column names for log-transformed columns
  data <- data[, !names(data) %in% log_columns]  # Remove log-transformed columns

  return(data)
}
```

```{r}
dataset_apartment_cleaned <- clean_dataset_with_log(dataset_apartment)
dataset_land_cleaned <- clean_dataset_with_log(dataset_land)
dataset_house_cleaned <- clean_dataset_with_log(dataset_house)

summary(dataset_apartment_cleaned)
summary(dataset_land_cleaned)
summary(dataset_house_cleaned)
```

This gave a much better result in the ones we had problems.

# 3. Main Question

**What characteristics should be prioritized to invest in houses in
Portugal?**

We divide the problem into 3 different but related sub questions, in
order to structure better the problem.

## First Subquestion

**Which internal factors have the greatest potential for revaluation?
What impact do additional services (parking, lift...) have on the
investment decision?**

```{r}
# Assign factor levels to the EnergyCertificate column for apartments
dataset_apartment_cleaned$EnergyCertificate <- factor(
  dataset_apartment_cleaned$EnergyCertificate,
  levels = c("No Certificate", "G", "F", "E", "D", "C", "B-", "B", "A", "A+")
)

# Assign factor levels to the EnergyCertificate column for houses
dataset_house_cleaned$EnergyCertificate <- factor(
  dataset_house_cleaned$EnergyCertificate,
  levels = c("No Certificate", "G", "F", "E", "D", "C", "B-", "B", "A", "A+")
)

# Assign factor levels to the EnergyCertificate column for land
dataset_land_cleaned$EnergyCertificate <- factor(
  dataset_land_cleaned$EnergyCertificate,
  levels = c("No Certificate", "G", "F", "E", "D", "C", "B-", "B", "A", "A+")
)

# Convert factors to numeric values for apartments
processed_data <- dataset_apartment_cleaned %>%
  mutate(across(where(is.factor), ~ as.numeric(as.factor(.))))  # Convert factors to numeric values

# Select numeric columns
all_numeric_vars <- processed_data %>% 
  select(where(is.numeric))

# Remove constant columns (zero variance)
all_numeric_vars <- all_numeric_vars %>%
  select(where(~ var(., na.rm = TRUE) > 0))

# Calculate the correlation matrix (handling missing values)
correlation_matrix <- cor(all_numeric_vars, use = "pairwise.complete.obs")

correlation_matrix

# Check that the correlation matrix does not contain NA/NaN values
if (any(is.na(correlation_matrix))) {
  stop("The correlation matrix contains NA/NaN. Please check the data.")
}

# Generate a heatmap for the correlation matrix
ggcorrplot(correlation_matrix, 
           hc.order = TRUE, 
           type = "lower", 
           lab = TRUE, 
           lab_size = 2, 
           colors = c("red", "white", "blue"),
           title = "Correlation Matrix Apartments",
           ggtheme = theme_minimal())

# Convert factors to numeric values for houses
processed_data <- dataset_house_cleaned %>%
  mutate(across(where(is.factor), ~ as.numeric(as.factor(.))))  # Convert factors to numeric values

# Select numeric columns
all_numeric_vars <- processed_data %>% 
  select(where(is.numeric))

# Remove constant columns (zero variance)
all_numeric_vars <- all_numeric_vars %>%
  select(where(~ var(., na.rm = TRUE) > 0))

# Calculate the correlation matrix (handling missing values)
correlation_matrix <- cor(all_numeric_vars, use = "pairwise.complete.obs")

correlation_matrix

# Check that the correlation matrix does not contain NA/NaN values
if (any(is.na(correlation_matrix))) {
  stop("The correlation matrix contains NA/NaN. Please check the data.")
}

# Generate a heatmap for the correlation matrix
ggcorrplot(correlation_matrix, 
           hc.order = TRUE, 
           type = "lower", 
           lab = TRUE, 
           lab_size = 2, 
           colors = c("red", "white", "blue"),
           title = "Correlation Matrix Houses",
           ggtheme = theme_minimal())

# Convert factors to numeric values for land
processed_data <- dataset_land_cleaned %>%
  mutate(across(where(is.factor), ~ as.numeric(as.factor(.))))  # Convert factors to numeric values

# Select numeric columns
all_numeric_vars <- processed_data %>% 
  select(where(is.numeric))

# Remove constant columns (zero variance)
all_numeric_vars <- all_numeric_vars %>%
  select(where(~ var(., na.rm = TRUE) > 0))

# Calculate the correlation matrix (handling missing values)
correlation_matrix <- cor(all_numeric_vars, use = "pairwise.complete.obs")

correlation_matrix

# Check that the correlation matrix does not contain NA/NaN values
if (any(is.na(correlation_matrix))) {
  stop("The correlation matrix contains NA/NaN. Please check the data.")
}

# Generate a heatmap for the correlation matrix
ggcorrplot(correlation_matrix, 
           hc.order = TRUE, 
           type = "lower", 
           lab = TRUE, 
           lab_size = 2, 
           colors = c("red", "white", "blue"),
           title = "Correlation Matrix Lands",
           ggtheme = theme_minimal())
```

Apartment, has great correlation with price: - NumberOfBathrooms
(0.404) - CombinedElevator (0.356) - Parking (0.328) - TotalArea
(0.310) - ConstructionYear (0.294) - LivingArea (0.286)

House, has great correlation with price: - NumberOfBathrooms: 0.632 -
EnergyCertificate: 0.530 - ConstructionYear: 0.511 - LivingArea: 0.453 -
Parking: 0.312 - TotalRooms: 0.317 - TotalArea: 0.268

Land, there are no significant correlations, so undervalued lands can be
evaluated based on price and area, as these are the only important
variables.

With these graphs, we can demonstrate that there is a correlation
between variables.

**Apartments**

```{r}
set.seed(42)
data <- dataset_apartment_cleaned

# Relationship 1: NumberOfBathrooms (categorical) vs Price (numerical) - Boxplot
ggplot(data, aes(x = as.factor(NumberOfBathrooms), y = Price)) +
  geom_boxplot(fill = "blue", alpha = 0.7) +
  ggtitle("Distribution of Price by Number of Bathrooms") +
  xlab("Number of Bathrooms") +
  ylab("Price")

# Relationship 2: Parking (categorical) vs Price (numerical) - Boxplot
ggplot(data, aes(x = as.factor(Parking), y = Price)) +
  geom_boxplot(fill = "green", alpha = 0.7) +
  ggtitle("Distribution of Price by Parking") +
  xlab("Parking") +
  ylab("Price")

# Relationship 3: TotalArea (numerical) vs Price (numerical) - Linear Regression
ggplot(data, aes(x = TotalArea, y = Price)) +
  geom_point(color = "purple") +
  geom_smooth(method = "lm", color = "red") +
  ggtitle("Total Area vs Price") +
  xlab("Total Area") +
  ylab("Price")

# Relationship 4: LivingArea (numerical) vs Price (numerical) - Linear Regression
ggplot(data, aes(x = LivingArea, y = Price)) +
  geom_point(color = "orange") +
  geom_smooth(method = "lm", color = "red") +
  ggtitle("Living Area vs Price") +
  xlab("Living Area") +
  ylab("Price")

# Relationship 5: ConstructionYear (numerical) vs Price (numerical) - Linear Regression
ggplot(data, aes(x = ConstructionYear, y = Price)) +
  geom_point(color = "cyan") +
  geom_smooth(method = "lm", color = "red") +
  ggtitle("Construction Year vs Price") +
  xlab("Construction Year") +
  ylab("Price")

# Relationship 6: CombinedElevator (categorical) vs Price (numerical) - Boxplot
ggplot(data, aes(x = as.factor(CombinedElevator), y = Price)) +
  geom_boxplot(fill = "pink", alpha = 0.7) +
  ggtitle("Combined Elevator vs Price") +
  xlab("Combined Elevator") +
  ylab("Price")
```

**Houses**

```{r}
set.seed(42)
data <- dataset_house_cleaned

# Relationship 1: NumberOfBathrooms (categorical) vs Price (Boxplot)
ggplot(data, aes(x = as.factor(NumberOfBathrooms), y = Price)) +
  geom_boxplot(fill = "blue", alpha = 0.7) +
  ggtitle("Price Distribution by Number of Bathrooms") +
  xlab("Number of Bathrooms") +
  ylab("Price")

# Relationship 2: EnergyCertificate (categorical) vs Price (Boxplot)
ggplot(data, aes(x = as.factor(EnergyCertificate), y = Price)) +
  geom_boxplot(fill = "green", alpha = 0.7) +
  ggtitle("Price Distribution by Energy Certificate") +
  xlab("Energy Certificate") +
  ylab("Price")

# Relationship 3: ConstructionYear (numeric) vs Price (Linear Regression)
ggplot(data, aes(x = ConstructionYear, y = Price)) +
  geom_point(color = "purple") +
  geom_smooth(method = "lm", color = "red") +
  ggtitle("Construction Year vs Price") +
  xlab("Construction Year") +
  ylab("Price")

# Relationship 4: LivingArea (numeric) vs Price (Linear Regression)
ggplot(data, aes(x = LivingArea, y = Price)) +
  geom_point(color = "orange") +
  geom_smooth(method = "lm", color = "red") +
  ggtitle("Living Area vs Price") +
  xlab("Living Area") +
  ylab("Price")

# Relationship 5: Parking (categorical) vs Price (Boxplot)
ggplot(data, aes(x = as.factor(Parking), y = Price)) +
  geom_boxplot(fill = "pink", alpha = 0.7) +
  ggtitle("Price Distribution by Parking") +
  xlab("Parking") +
  ylab("Price")

# Relationship 6: TotalRooms (discrete numeric) vs Price (Boxplot)
ggplot(data, aes(x = as.factor(TotalRooms), y = Price)) +
  geom_boxplot(fill = "cyan", alpha = 0.7) +
  ggtitle("Price Distribution by Total Rooms") +
  xlab("Total Rooms") +
  ylab("Price")

# Relationship 7: TotalArea (numeric) vs Price (Linear Regression)
ggplot(data, aes(x = TotalArea, y = Price)) +
  geom_point(color = "darkgreen") +
  geom_smooth(method = "lm", color = "red") +
  ggtitle("Total Area vs Price") +
  xlab("Total Area") +
  ylab("Price")
```

**Lands**

```{r}
set.seed(42)
data <- dataset_land_cleaned

# Relationship 1: EnergyCertificate (categorical) vs Price (Boxplot)
ggplot(data, aes(x = EnergyCertificate, y = Price)) +
  geom_boxplot(fill = "purple", alpha = 0.7) +
  ggtitle("Price Distribution by Energy Certificate") +
  xlab("Energy Certificate") +
  ylab("Price")

# Relationship 2: TotalArea (numeric) vs Price (Linear Regression)
ggplot(data, aes(x = TotalArea, y = Price)) +
  geom_point(color = "cyan") +
  geom_smooth(method = "lm", color = "red") +
  ggtitle("Total Area vs Price") +
  xlab("Total Area") +
  ylab("Price")

# Relationship 3: CombinedElevator (numeric) vs Price (Linear Regression)
ggplot(data, aes(x = as.factor(CombinedElevator), y = Price)) +
  geom_boxplot(fill = "pink", alpha = 0.7) +
  ggtitle("Combined Elevator vs Price") +
  xlab("Combined Elevator") +
  ylab("Price")
```

We can see how Land has almost no correlation with any variable.

In the second section, the characteristics of homes with a high price
will be estimated, and a search will be made for those homes with
similar properties and a lower price.

## Second Subquestion

**What are the common characteristics of homes with high prices? How can
we identify similar residences that are undervalued?**

### First Method

**Apartments**

For the previous study, we know that the characteristics most related to
the price of apartments are:

Apartment, high correlation with the price: - NumberOfBathrooms
(0.404) - CombinedElevator (0.356) - Parking (0.328) - TotalArea
(0.310) - ConstructionYear (0.294) - LivingArea (0.286)

```{r}
# 2. Initial Predictive Model
price_model <- lm(Price ~ TotalArea + LivingArea + Parking + 
                  ConstructionYear + NumberOfBathrooms + CombinedElevator,
                  data = dataset_apartment_cleaned)

# 3. Prediction and residuals calculation
dataset_apartment_cleaned <- dataset_apartment_cleaned %>%
  mutate(
    Predicted_Price = predict(price_model, dataset_apartment_cleaned),
    Predicted_Error = residuals(price_model), # Model errors
    Lower_Bound = Predicted_Price - 1.96 * sd(Predicted_Error), # Lower bound
    Upper_Bound = Predicted_Price + 1.96 * sd(Predicted_Error)  # Upper bound
  )

# 4. Refine undervalued definition with confidence intervals
dataset_apartment_cleaned <- dataset_apartment_cleaned %>%
  mutate(
    Price_Category = case_when(
      Price < Lower_Bound ~ "Undervalued",
      Price > Upper_Bound ~ "Overvalued",
      TRUE ~ "Fairly Valued"
    )
  )

# 5. Compare actual and predicted prices with ratios
dataset_apartment_cleaned <- dataset_apartment_cleaned %>%
  mutate(
    Price_Ratio = Price / Predicted_Price,
    Refined_Price_Category = case_when(
      Price_Ratio < 0.8 ~ "Undervalued",
      Price_Ratio > 1.2 ~ "Overvalued",
      TRUE ~ "Fairly Valued"
    )
  )

# 6. Clustering analysis by similar characteristics
cluster_data <- dataset_apartment_cleaned %>%
  select(TotalArea, LivingArea, TotalRooms, NumberOfBathrooms, ConstructionYear, Parking) %>%
  scale()

set.seed(123)
kmeans_result <- kmeans(cluster_data, centers = 4)

# Add clusters to the main dataset
dataset_apartment_cleaned <- dataset_apartment_cleaned %>%
  mutate(Cluster = kmeans_result$cluster)

# Calculate average prices by cluster
cluster_summary <- dataset_apartment_cleaned %>%
  group_by(Cluster) %>%
  summarise(
    Avg_Price = mean(Price, na.rm = TRUE),
    Avg_Predicted = mean(Predicted_Price, na.rm = TRUE)
  )

# Ensure the join is correct
dataset_apartment_cleaned <- dataset_apartment_cleaned %>%
  left_join(cluster_summary, by = "Cluster")

# Validate that necessary columns are present
if (!"Avg_Price" %in% names(dataset_apartment_cleaned)) {
  stop("Error: 'Avg_Price' was not correctly added to the main dataset.")
}

# Identify undervalued apartments by cluster
dataset_apartment_cleaned <- dataset_apartment_cleaned %>%
  mutate(
    Cluster_Price_Category = case_when(
      Price < 0.8 * Avg_Price ~ "Undervalued",
      Price > 1.2 * Avg_Price ~ "Overvalued",
      TRUE ~ "Fairly Valued"
    )
  )

# 7. Visualizations
# Bar plot of the distribution of categories
# Proportion of categories within each cluster
dataset_apartment_cleaned %>%
  group_by(Cluster, Cluster_Price_Category) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100) %>%
  ggplot(aes(x = as.factor(Cluster), y = Percentage, fill = Cluster_Price_Category)) +
  geom_bar(stat = "identity", position = "fill") +
  ggtitle("Proportion of Categories by Cluster") +
  xlab("Cluster") +
  ylab("Proportion (%)") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red", "green"))


ggplot(dataset_apartment_cleaned, aes(x = Price_Ratio, fill = Refined_Price_Category)) +
  geom_histogram(binwidth = 0.05, alpha = 0.7) +
  ggtitle("Distribution of Price Ratios by Category") +
  theme_minimal()

ggplot(dataset_apartment_cleaned, aes(x = Cluster, y = Price, color = Cluster_Price_Category)) +
  geom_boxplot() +
  ggtitle("Price Categories by Cluster") +
  theme_minimal()

ggplot(dataset_apartment_cleaned, aes(x = TotalArea, y = Price - Predicted_Price, color = Price_Category)) +
  geom_point(alpha = 0.7) +
  ggtitle("Refined Price Categories with Confidence Intervals") +
  theme_minimal()

ggplot(dataset_apartment_cleaned, aes(x = as.numeric(CombinedElevator), y = Price - Predicted_Price, color = Price_Category)) +
  geom_boxplot() +
  ggtitle("Elevator Availability vs Price Difference") +
  theme_minimal()

# 8. Final summary of results
summary_stats <- dataset_apartment_cleaned %>%
  group_by(Cluster_Price_Category) %>%
  summarise(
    Avg_TotalArea = mean(TotalArea, na.rm = TRUE),
    Avg_LivingArea = mean(LivingArea, na.rm = TRUE),
    Avg_TotalRooms = mean(TotalRooms, na.rm = TRUE),
    Avg_NumberOfBathrooms = mean(NumberOfBathrooms, na.rm = TRUE),
    Avg_ConstructionYear = mean(ConstructionYear, na.rm = TRUE),
    Avg_Parking = mean(Parking, na.rm = TRUE),
    Count = n()
  )

print(summary_stats)
```

**Interpretation of Results**

1.  **Correlation Matrix**

-   TotalArea and LivingArea have a high correlation with price (Price),
    confirming that the size of the apartment is one of the most
    influential factors.
-   NumberOfBathrooms and ConstructionYear also show a positive
    correlation, indicating that newer apartments with more bathrooms
    tend to be more expensive.
-   Parking has a low but positive correlation, suggesting it adds value
    but is not as decisive as area or age.

2.  **Distribution of Ratios (Actual Price / Predicted)**

-   The Undervalued category appears in the lowest ratio range (\< 0.8).
-   Overvalued apartments are concentrated in ratios above 1.2.
-   The majority of apartments are Fairly Valued, with ratios close to
    1.  

3.  **Categories by Cluster**

-   Clusters help group apartments with similar characteristics (such as
    area, number of rooms and bathrooms, etc.).
-   Undervalued apartments appear in clusters with lower average prices,
    indicating they have similar characteristics to others in the group,
    but their actual price is much lower.

4.  **Refining Categories with Confidence Intervals**

-   The blue points (Undervalued) in the "Refined Price Categories with
    Confidence Intervals" graph are concentrated at the lower end,
    showing that undervalued apartments have a significantly lower
    actual price than expected.
-   The majority of apartments are fairly valued (Fairly Valued), and
    overvalued apartments (Overvalued) appear scattered at the upper
    end.

5.  **Impact of Elevators on Price**

-   Apartments with elevators (CombinedElevator) tend to be more highly
    valued or overvalued.
-   Apartments without elevators are more likely to be undervalued.

6.  **Summary by Category (Table)**

-   **Overvalued:** Apartments with more area, more bathrooms, and built
    more recently (average year: 2001).
-   **Fairly Valued:** Balanced area and characteristics compared to
    other categories.
-   **Undervalued:** Apartments with slightly smaller total and living
    area and an average construction year lower (1998).

------------------------------------------------------------------------

**Conclusions (so far)**

**Factors Determining High Prices:**

-   Larger total and living area.
-   More bathrooms and modern features.
-   Elevator and parking availability.

**Identification of Undervalued Apartments:**

-   Lower ratio between actual and predicted price (ratios \< 0.8).
-   Apartments in clusters with lower average prices.
-   Absence of elevator and less modern features.

**Key Differentiators:**

-   Construction year (newer apartments have higher prices).
-   Total number of bathrooms and living area.

**Lands**

We found no significant correlations between the available variables and
the price of lands. Therefore, for the analysis, we will focus on the
most relevant variables: **TotalArea** (the size of the land) and
**City** (location), as these are likely to have the greatest impact on
price based on domain knowledge.

The process is going to be similar than in the study with
apartments.But, as we have no correlations with price, we will use a
non-linear model. In the following process we are looking for the best
one.

```{r}
# Convert 'City' variable into numeric to treat it as a categorical variable
dataset_land_cleaned <- dataset_land_cleaned %>%
  mutate(City = as.numeric(as.factor(City)))  # Ensure City is a categorical variable
```

We fit three different types of models:

-   **Polynomial Model**: A polynomial regression model that can capture
    non-linear relationships.
-   **Exponential Model**: A model that assumes an exponential
    relationship by using the log of the target variable (Price).
-   **Power Model**: A model using a power transformation
    (TotalArea\^0.5) to capture potential non-linearities in TotalArea.

```{r}
# Fit a Polynomial model
price_model_poly <- lm(Price ~ poly(TotalArea, 2) + poly(City, 2), data = dataset_land_cleaned)

# Fit an Exponential model by taking the log of the Price
price_model_expo <- lm(log(Price) ~ TotalArea + City, data = dataset_land_cleaned)

# Fit a Power model by applying a power transformation to TotalArea
price_model_power <- lm(Price ~ I(TotalArea^0.5) + City, data = dataset_land_cleaned)
```

After fitting the models, we evaluate their performance using two
metrics:

-   **R-squared (R²)**: This indicates the proportion of variance
    explained by the model. Higher R² values indicate a better fit.

-   **Root Mean Square Error (RMSE)**: This measures the average
    deviation between the predicted and actual values. Lower RMSE
    indicates a better fit.

```{r}
# Evaluate performance for each model using R-squared and RMSE

# 3.1. Polynomial Model
dataset_land_cleaned <- dataset_land_cleaned %>%
  mutate(Predicted_Price_Poly = predict(price_model_poly, dataset_land_cleaned))

R2_poly <- summary(price_model_poly)$r.squared  # R-squared for the polynomial model
RMSE_poly <- sqrt(mean((dataset_land_cleaned$Price - dataset_land_cleaned$Predicted_Price_Poly)^2))  # RMSE for polynomial model

# 3.2. Exponential Model
dataset_land_cleaned <- dataset_land_cleaned %>%
  mutate(Predicted_Price_Expo = exp(predict(price_model_expo, dataset_land_cleaned)))

R2_expo <- summary(price_model_expo)$r.squared  # R-squared for the exponential model
RMSE_expo <- sqrt(mean((dataset_land_cleaned$Price - dataset_land_cleaned$Predicted_Price_Expo)^2))  # RMSE for exponential model

# 3.3. Power Model
dataset_land_cleaned <- dataset_land_cleaned %>%
  mutate(Predicted_Price_Power = predict(price_model_power, dataset_land_cleaned))

R2_power <- summary(price_model_power)$r.squared  # R-squared for the power model
RMSE_power <- sqrt(mean((dataset_land_cleaned$Price - dataset_land_cleaned$Predicted_Price_Power)^2))  # RMSE for power model
```

A comparison table is created to assess which model performs the best.
The models are ranked based on R² and RMSE, with the highest R² and
lowest RMSE indicating the best fit.

```{r}
# Create a summary table to compare R-squared and RMSE for each model
model_comparison <- data.frame(
  Model = c("Polynomial", "Exponential", "Power"),
  R2 = c(R2_poly, R2_expo, R2_power),
  RMSE = c(RMSE_poly, RMSE_expo, RMSE_power)
)

# Print model comparison
print(model_comparison)

# The best model is selected based on the highest R-squared and lowest RMSE
best_model <- model_comparison %>%
  arrange(desc(R2), RMSE) %>%
  slice(1)  # Select the top model based on the criteria

# Print the name of the best model
print(paste("The best model is:", best_model$Model))
```

The best model is selected based on the highest R² and the lowest RMSE.
This model is **Exponential**.

```{r}
summary(dataset_land_cleaned)
```

```{r}
# 1. Convert 'City' to a categorical variable (One-Hot Encoding)
dataset_land_cleaned <- dataset_land_cleaned %>%
  mutate(City = as.numeric(as.factor(City)))  # Ensure 'City' is treated as a categorical variable

summary(dataset_land_cleaned)

# 2. Exponential Predictive Model
# Using 'TotalArea' and 'City' as predictors, with the log-transformed 'Price' for the exponential model
price_model_expo <- lm(log(Price) ~ TotalArea + City, data = dataset_land_cleaned)

# 3. Prediction and residual calculation
dataset_land_cleaned <- dataset_land_cleaned %>%
  mutate(
    Predicted_Price = exp(predict(price_model_expo, dataset_land_cleaned)),  # Applying exp to get predicted price
    Predicted_Error = Price - Predicted_Price, # Residuals (difference between actual and predicted price)
    Lower_Bound = Predicted_Price - 1.96 * sd(Predicted_Error), # Lower bound for confidence intervals
    Upper_Bound = Predicted_Price + 1.96 * sd(Predicted_Error)  # Upper bound for confidence intervals
  )

# 4. Refine undervaluation definition using confidence intervals
dataset_land_cleaned <- dataset_land_cleaned %>%
  mutate(
    Price_Category = case_when(
      Price < Lower_Bound ~ "Undervalued",
      Price > Upper_Bound ~ "Overvalued",
      TRUE ~ "Fairly Valued"
    )
  )

# 5. Compare actual and predicted prices using ratios
dataset_land_cleaned <- dataset_land_cleaned %>%
  mutate(
    Price_Ratio = Price / Predicted_Price,
    Refined_Price_Category = case_when(
      Price_Ratio < 0.8 ~ "Undervalued",
      Price_Ratio > 1.2 ~ "Overvalued",
      TRUE ~ "Fairly Valued"
    )
  )

# 6. Clustering analysis by similar characteristics
cluster_data <- dataset_land_cleaned %>%
  select(TotalArea, City) %>%
  scale()  # Standardize data before clustering

set.seed(123)
kmeans_result <- kmeans(cluster_data, centers = 4)

# Add clusters to the main dataset
dataset_land_cleaned <- dataset_land_cleaned %>%
  mutate(Cluster = kmeans_result$cluster)

# Calculate average prices by cluster
cluster_summary <- dataset_land_cleaned %>%
  group_by(Cluster) %>%
  summarise(
    Avg_Price = mean(Price, na.rm = TRUE),
    Avg_Predicted = mean(Predicted_Price, na.rm = TRUE)
  )

# Ensure correct merging of summary with the main dataset
dataset_land_cleaned <- dataset_land_cleaned %>%
  left_join(cluster_summary, by = "Cluster")

# Validate that the necessary columns are present
if (!"Avg_Price" %in% names(dataset_land_cleaned)) {
  stop("Error: 'Avg_Price' was not correctly added to the main dataset.")
}

# Identify undervalued apartments by cluster
dataset_land_cleaned <- dataset_land_cleaned %>%
  mutate(
    Cluster_Price_Category = case_when(
      Price < 0.8 * Avg_Price ~ "Undervalued",
      Price > 1.2 * Avg_Price ~ "Overvalued",
      TRUE ~ "Fairly Valued"
    )
  )

# 7. Visualizations
# Bar plot for distribution of categories by cluster
dataset_land_cleaned %>%
  group_by(Cluster, Cluster_Price_Category) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100) %>%
  ggplot(aes(x = as.factor(Cluster), y = Percentage, fill = Cluster_Price_Category)) +
  geom_bar(stat = "identity", position = "fill") +
  ggtitle("Proportion of Categories by Cluster") +
  xlab("Cluster") +
  ylab("Proportion (%)") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red", "green"))

ggplot(dataset_land_cleaned, aes(x = Price_Ratio, fill = Refined_Price_Category)) +
  geom_histogram(binwidth = 0.05, alpha = 0.7) +
  ggtitle("Distribution of Price Ratios by Category") +
  theme_minimal()

ggplot(dataset_land_cleaned, aes(x = Cluster, y = Price, color = Cluster_Price_Category)) +
  geom_boxplot() +
  ggtitle("Price Categories by Cluster") +
  theme_minimal()

ggplot(dataset_land_cleaned, aes(x = TotalArea, y = Price - Predicted_Price, color = Price_Category)) +
  geom_point(alpha = 0.7) +
  ggtitle("Refined Price Categories with Confidence Intervals") +
  theme_minimal()

ggplot(dataset_land_cleaned, aes(x = as.numeric(City), y = Price - Predicted_Price, color = Price_Category)) +
  geom_boxplot() +
  ggtitle("City vs Price Difference") +
  theme_minimal()

# 8. Final summary of results
summary_stats <- dataset_land_cleaned %>%
  group_by(Cluster_Price_Category) %>%
  summarise(
    Avg_TotalArea = mean(TotalArea, na.rm = TRUE),
    Avg_City = mean(City, na.rm = TRUE),
    Count = n()
  )

print(summary_stats)
```

**Houses**

The same proces as before is going to be used. The variables with higher
correlation with price are:

-   NumberOfBathrooms (0.63)
-   EnergyCertificate (0.53)
-   ConstructionYear (0.51)
-   LivingArea (0.45)

```{r}
summary(dataset_house_cleaned)
```

```{r}
# 1. Transform EnergyCertificate to numeric.
dataset_house_cleaned <- dataset_house_cleaned %>%
  mutate(EnergyCertificate = as.numeric(EnergyCertificate))

# Transform Price using natural logarithm
dataset_house_cleaned <- dataset_house_cleaned %>%
  mutate(Log_Price = log(Price))

# Fit the model with the transformed variable
price_model <- lm(Log_Price ~ NumberOfBathrooms + LivingArea + 
                  ConstructionYear + EnergyCertificate,
                  data = dataset_house_cleaned)

# 3. Prediction and calculation of residuals
dataset_house_cleaned <- dataset_house_cleaned %>%
  mutate(
    Predicted_Price = predict(price_model, dataset_house_cleaned),
    Predicted_Error = residuals(price_model), # Model errors
    Lower_Bound = Predicted_Price - 1.96 * sd(Predicted_Error), # Lower bound
    Upper_Bound = Predicted_Price + 1.96 * sd(Predicted_Error)  # Upper bound
  )

# 4. Refine definition of undervalued with confidence intervals
dataset_house_cleaned <- dataset_house_cleaned %>%
  mutate(
    Price_Category = case_when(
      Log_Price < Lower_Bound ~ "Undervalued",
      Log_Price > Upper_Bound ~ "Overvalued",
      TRUE ~ "Fairly Valued"
    )
  )

# 5. Compare actual and predicted prices with ratios
dataset_house_cleaned <- dataset_house_cleaned %>%
  mutate(
    Price_Ratio = Log_Price / Predicted_Price,
    Refined_Price_Category = case_when(
      Price_Ratio < 0.95 ~ "Undervalued",
      Price_Ratio > 1.05 ~ "Overvalued",
      TRUE ~ "Fairly Valued"
    )
  )
```

```{r}
# Filter records with Price less than 0
negative_price_samples <- dataset_house_cleaned %>%
  filter(Predicted_Price < 0)

# Print the filtered records
print(negative_price_samples)

# Optional: Check how many records meet this condition
cat("Number of records with Price < 0:", nrow(negative_price_samples), "\n")
```

```{r}
# 6. Clustering analysis by similar characteristics
cluster_data <- dataset_house_cleaned %>%
  select(LivingArea, EnergyCertificate, NumberOfBathrooms, ConstructionYear) %>%
  scale()

set.seed(123)
kmeans_result <- kmeans(cluster_data, centers = 4)

# Add the clusters to the main dataset
dataset_house_cleaned <- dataset_house_cleaned %>%
  mutate(Cluster = kmeans_result$cluster)

# Calculate average prices by cluster
cluster_summary <- dataset_house_cleaned %>%
  group_by(Cluster) %>%
  summarise(
    Avg_Price = mean(Log_Price, na.rm = TRUE),
    Avg_Predicted = mean(Predicted_Price, na.rm = TRUE)
  )

# Ensure the join is correct
dataset_house_cleaned <- dataset_house_cleaned %>%
  left_join(cluster_summary, by = "Cluster")

# Validate that the necessary columns are present
if (!"Avg_Price" %in% names(dataset_house_cleaned)) {
  stop("Error: 'Avg_Price' was not correctly added to the main dataset.")
}

# Identify undervalued houses by cluster
dataset_house_cleaned <- dataset_house_cleaned %>%
  mutate(
    Cluster_Price_Category = case_when(
      Log_Price < 0.95 * Avg_Price ~ "Undervalued",
      Log_Price > 1.05 * Avg_Price ~ "Overvalued",
      TRUE ~ "Fairly Valued"
    )
  )

# 7. Visualizations
# Bar chart of the distribution of categories
# Proportion of categories within each cluster
dataset_house_cleaned %>%
  group_by(Cluster, Cluster_Price_Category) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100) %>%
  ggplot(aes(x = as.factor(Cluster), y = Percentage, fill = Cluster_Price_Category)) +
  geom_bar(stat = "identity", position = "fill") +
  ggtitle("Proportion of Categories by Cluster") +
  xlab("Cluster") +
  ylab("Proportion (%)") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red", "green"))
```

```{r}
ggplot(dataset_house_cleaned, aes(x = Price_Ratio, fill = Refined_Price_Category)) +
  geom_histogram(binwidth = 0.05, alpha = 0.7) +
  ggtitle("Distribution of Price Ratios by Category") +
  theme_minimal()

ggplot(dataset_house_cleaned, aes(x = Cluster, y = Log_Price, color = Cluster_Price_Category)) +
  geom_boxplot() +
  ggtitle("Price Categories by Cluster") +
  theme_minimal()

ggplot(dataset_house_cleaned, aes(x = TotalArea, y = Log_Price - Predicted_Price, color = Price_Category)) +
  geom_point(alpha = 0.7) +
  ggtitle("Refined Price Categories with Confidence Intervals") +
  theme_minimal()

ggplot(dataset_house_cleaned, aes(x = as.numeric(NumberOfBathrooms), y = Log_Price - Predicted_Price, color = Price_Category)) +
  geom_boxplot() +
  ggtitle("Number of Bathrooms vs Price Difference") +
  theme_minimal()
```

```{r}
# 8. Summary of final results
summary_stats <- dataset_house_cleaned %>%
  group_by(Cluster_Price_Category) %>%
  summarise(
    Avg_TotalArea = mean(TotalArea, na.rm = TRUE),
    Avg_EnergyCertificate = mean(EnergyCertificate, na.rm = TRUE),
    Avg_NumberOfBathrooms = mean(NumberOfBathrooms, na.rm = TRUE),
    Avg_ConstructionYear = mean(ConstructionYear, na.rm = TRUE),
    Count = n()
  )

print(summary_stats)
```

### Second Method

Now we are going to make another process to find overvalued and
undervalued properties.

This process identifies overvalued and undervalued properties by
clustering them based on their more influent characteristics (excluding
price) and then comparing prices within each cluster. Here's what
happens:

1.  Clustering Without Price: Properties are grouped into clusters using
    the attributes that hace more correlation with price in each
    dataset. Price is not included in clustering to avoid bias.

2.  Price-Based Classification: For each cluster, the average price is
    calculated. Properties with prices:

-   Below 80% of the cluster average are labeled undervalued.
-   Above 120% of the cluster average are labeled overvalued.
-   Within 80%-120% are labeled fairly valued.

3.  Visualizations: Charts display the proportion of categories
    (undervalued, overvalued, fairly valued) and the price distribution
    within each cluster.

This approach separates grouping properties by features from evaluating
their prices, providing an unbiased perspective on valuation.

**Apartments**

```{r}
# 1. Prepare data for clustering (excluding price)
cluster_data <- dataset_apartment_cleaned %>%
  select(TotalArea, LivingArea, TotalRooms, NumberOfBathrooms, ConstructionYear, Parking) %>%
  scale() # Scale the data for clustering

# 2. Perform k-means clustering
set.seed(123)
kmeans_result <- kmeans(cluster_data, centers = 4) # Use 4 clusters (adjust as needed)

# Add cluster assignments to the dataset
dataset_apartment_cleaned <- dataset_apartment_cleaned %>%
  mutate(Cluster = kmeans_result$cluster)

# 3. Calculate average price and classify properties within each cluster
dataset_apartment_cleaned <- dataset_apartment_cleaned %>%
  group_by(Cluster) %>%
  mutate(
    Avg_Price_Cluster = mean(Price, na.rm = TRUE),
    Price_Category_Cluster = case_when(
      Price < 0.8 * Avg_Price_Cluster ~ "Undervalued",  # Less than 80% of the cluster average
      Price > 1.2 * Avg_Price_Cluster ~ "Overvalued",  # Greater than 120% of the cluster average
      TRUE ~ "Fairly Valued"  # Within 80%-120% of the cluster average
    )
  ) %>%
  ungroup() # Ungroup after processing

# 4. Summarize results: Count of categories within each cluster
cluster_summary <- dataset_apartment_cleaned %>%
  group_by(Cluster, Price_Category_Cluster) %>%
  summarise(Count = n(), .groups = "drop")

# Print the cluster summary
print(cluster_summary)

# 5. Visualizations
# Proportion of categories within each cluster
ggplot(cluster_summary, aes(x = as.factor(Cluster), y = Count, fill = Price_Category_Cluster)) +
  geom_bar(stat = "identity", position = "fill") +
  ggtitle("Proportion of Price Categories Within Each Cluster") +
  xlab("Cluster") +
  ylab("Proportion") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red", "green"))

# Boxplot of prices within each cluster
ggplot(dataset_apartment_cleaned, aes(x = as.factor(Cluster), y = Price, color = Price_Category_Cluster)) +
  geom_boxplot() +
  ggtitle("Price Distribution Within Each Cluster") +
  xlab("Cluster") +
  ylab("Price") +
  theme_minimal()

```

The third cluster stands out as the most interesting because it shows a
balanced distribution of undervalued, fairly valued, and overvalued
properties. Additionally, as shown in the second plot, this cluster has
a higher number of undervalued properties and a greater quantity of
low-priced homes compared to the average for this cluster.

```{r}
# Filter for undervalued houses in the third cluster
undervalued_cluster_3 <- dataset_apartment_cleaned %>%
  filter(Cluster == 3, Price_Category_Cluster == "Undervalued")

# Summarize common characteristics
summary_undervalued_cluster_3 <- undervalued_cluster_3 %>%
  summarise(
    Avg_TotalArea = mean(TotalArea, na.rm = TRUE),
    Median_TotalArea = median(TotalArea, na.rm = TRUE),
    Avg_LivingArea = mean(LivingArea, na.rm = TRUE),
    Median_LivingArea = median(LivingArea, na.rm = TRUE),
    Avg_TotalRooms = mean(TotalRooms, na.rm = TRUE),
    Median_TotalRooms = median(TotalRooms, na.rm = TRUE),
    Avg_NumberOfBathrooms = mean(NumberOfBathrooms, na.rm = TRUE),
    Median_NumberOfBathrooms = median(NumberOfBathrooms, na.rm = TRUE),
    Avg_ConstructionYear = mean(ConstructionYear, na.rm = TRUE),
    Median_ConstructionYear = median(ConstructionYear, na.rm = TRUE),
    Avg_Parking = mean(Parking, na.rm = TRUE),
    Median_Parking = median(Parking, na.rm = TRUE)
  )

# Print the summary
print(summary_undervalued_cluster_3)

# Print a few undervalued samples from this cluster
cat("\nSome undervalued properties in Cluster 3:\n")
print(head(undervalued_cluster_3, 50))  # Adjust the number of rows to display as needed

```

**Lands**

```{r}
# 1. Transform City and District to numeric without removing the original columns
dataset_land_cleaned <- dataset_land_cleaned %>%
  mutate(
    CityNum = as.numeric(City),  # Convert cities to numeric
    DistrictNum = as.numeric(District)  # Convert districts to numeric
  )

# 2. Prepare data for clustering (excluding price)
cluster_data <- dataset_land_cleaned %>%
  select(TotalArea, DistrictNum, CityNum) %>%
  scale() # Scale the data for clustering

# 3. Perform k-means clustering
set.seed(123)
kmeans_result <- kmeans(cluster_data, centers = 4)  # Use 4 clusters, you can adjust the number

# 4. Add cluster assignments to the dataset
dataset_land_cleaned <- dataset_land_cleaned %>%
  mutate(Cluster = kmeans_result$cluster)

# 5. Calculate average price and classify properties within each cluster
dataset_land_cleaned <- dataset_land_cleaned %>%
  group_by(Cluster) %>%
  mutate(
    Avg_Price_Cluster = mean(Price, na.rm = TRUE),
    Price_Category_Cluster = case_when(
      Price < 0.8 * Avg_Price_Cluster ~ "Undervalued",  # Less than 80% of the cluster average
      Price > 1.2 * Avg_Price_Cluster ~ "Overvalued",  # More than 120% of the cluster average
      TRUE ~ "Fairly Valued"  # Within 80%-120% of the cluster average
    )
  ) %>%
  ungroup()  # Ungroup after processing

# 6. Summarize results: Count of categories within each cluster
cluster_summary <- dataset_land_cleaned %>%
  group_by(Cluster, Price_Category_Cluster) %>%
  summarise(Count = n(), .groups = "drop")

# Print the cluster summary
print(cluster_summary)

# 7. Visualization: Proportion of categories within each cluster
ggplot(cluster_summary, aes(x = as.factor(Cluster), y = Count, fill = Price_Category_Cluster)) +
  geom_bar(stat = "identity", position = "fill") +
  ggtitle("Proportion of Price Categories Within Each Cluster") +
  xlab("Cluster") +
  ylab("Proportion") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red", "green"))

# 8. Boxplot of prices within each cluster
ggplot(dataset_land_cleaned, aes(x = as.factor(Cluster), y = Price, color = Price_Category_Cluster)) +
  geom_boxplot() +
  ggtitle("Price Distribution Within Each Cluster") +
  xlab("Cluster") +
  ylab("Price") +
  theme_minimal()
```

Although all the clusters are quite similar, the second cluster stands
out the most. It shows a balanced distribution of undervalued, fairly
valued, and overvalued properties. Additionally, as shown in the second
plot, this cluster has a higher number of undervalued properties and a
greater quantity of low-priced homes compared to the average for this
cluster.

```{r}
# 9. Filter undervalued properties in the second cluster
undervalued_cluster_2 <- dataset_land_cleaned %>%
  filter(Cluster == 2, Price_Category_Cluster == "Undervalued")

# 10. Summarize common characteristics of undervalued properties
summary_undervalued_cluster_2 <- undervalued_cluster_2 %>%
  summarise(
    Avg_TotalArea = mean(TotalArea, na.rm = TRUE),
    Median_TotalArea = median(TotalArea, na.rm = TRUE),
    Avg_District = mean(DistrictNum, na.rm = TRUE),
    Median_District = median(DistrictNum, na.rm = TRUE),
    Avg_City = mean(CityNum, na.rm = TRUE),
    Median_City = median(CityNum, na.rm = TRUE)
  )

# Print the summary
print(summary_undervalued_cluster_2)

# 11. Print some undervalued properties from cluster 2
cat("\nSome undervalued properties in Cluster 2:\n")
print(head(undervalued_cluster_2, 50))  # Adjust the number of rows to display as needed

```

**House**

```{r}
# 1. Transformamos EnergyCertificate a numerico.
dataset_house_cleaned <- dataset_house_cleaned %>%
  mutate(EnergyCertificate = as.numeric(EnergyCertificate))

# 1. Prepare data for clustering (excluding price)
cluster_data <- dataset_house_cleaned %>%
  select(LivingArea, ConstructionYear, EnergyCertificate, NumberOfBathrooms) %>%
  scale() # Scale the data for clustering

# 2. Perform k-means clustering
set.seed(123)
kmeans_result <- kmeans(cluster_data, centers = 4) # Use 4 clusters (adjust as needed)

# Add cluster assignments to the dataset
dataset_house_cleaned <- dataset_house_cleaned %>%
  mutate(Cluster = kmeans_result$cluster)

# 3. Calculate average price and classify properties within each cluster
dataset_house_cleaned <- dataset_house_cleaned %>%
  group_by(Cluster) %>%
  mutate(
    Avg_Price_Cluster = mean(Price, na.rm = TRUE),
    Price_Category_Cluster = case_when(
      Price < 0.8 * Avg_Price_Cluster ~ "Undervalued",  # Less than 80% of the cluster average
      Price > 1.2 * Avg_Price_Cluster ~ "Overvalued",  # Greater than 120% of the cluster average
      TRUE ~ "Fairly Valued"  # Within 80%-120% of the cluster average
    )
  ) %>%
  ungroup() # Ungroup after processing

# 4. Summarize results: Count of categories within each cluster
cluster_summary <- dataset_house_cleaned %>%
  group_by(Cluster, Price_Category_Cluster) %>%
  summarise(Count = n(), .groups = "drop")

# Print the cluster summary
print(cluster_summary)

# 5. Visualizations
# Proportion of categories within each cluster
ggplot(cluster_summary, aes(x = as.factor(Cluster), y = Count, fill = Price_Category_Cluster)) +
  geom_bar(stat = "identity", position = "fill") +
  ggtitle("Proportion of Price Categories Within Each Cluster") +
  xlab("Cluster") +
  ylab("Proportion") +
  theme_minimal() +
  scale_fill_manual(values = c("blue", "red", "green"))

# Boxplot of prices within each cluster
ggplot(dataset_house_cleaned, aes(x = as.factor(Cluster), y = Price, color = Price_Category_Cluster)) +
  geom_boxplot() +
  ggtitle("Price Distribution Within Each Cluster") +
  xlab("Cluster") +
  ylab("Price") +
  theme_minimal()

```

We can see that the second cluster is the most interesting, because it
has a good distribution of Undervalued, Fairly and Overvalued; and also,
as is shown in the second plot, it has more undervalued properties and
with a high quantity of low-price house in comparative with the mean of
this cluster.

```{r}
# Filter for undervalued houses in the second cluster
undervalued_cluster_2 <- dataset_house_cleaned %>%
  filter(Cluster == 2, Price_Category_Cluster == "Undervalued")

# Summarize common characteristics
summary_undervalued_cluster_2 <- undervalued_cluster_2 %>%
  summarise(
    Avg_LivingArea = mean(LivingArea, na.rm = TRUE),
    Median_LivingArea = median(LivingArea, na.rm = TRUE),
    Avg_ConstructionYear = mean(ConstructionYear, na.rm = TRUE),
    Median_ConstructionYear = median(ConstructionYear, na.rm = TRUE),
    Avg_EnergyCertificate = mean(EnergyCertificate, na.rm = TRUE),
    Median_EnergyCertificate = median(EnergyCertificate, na.rm = TRUE),
    Avg_NumberOfBathrooms = mean(NumberOfBathrooms, na.rm = TRUE),
    Median_NumberOfBathrooms = median(NumberOfBathrooms, na.rm = TRUE)
  )

# Print the summary
print(summary_undervalued_cluster_2)

# Print a few undervalued samples from this cluster
cat("\nSome undervalued properties in Cluster 2:\n")
print(head(undervalued_cluster_2, 50))  # Adjust the number of rows to display as needed

```

## Third Subquestion

**Which areas have the highest concentration of undervalued
properties?**

```{r}
undervalued_apartments <- dataset_apartment_cleaned %>%
  filter(Price_Category_Cluster == "Undervalued")

# Count the total properties by district
total_apartment_district <- dataset_apartment_cleaned %>%
  group_by(District) %>%
  summarise(Total = n(), .groups = "drop")

# Count the undervalued properties by district
undervalued_apartments_district <- undervalued_apartments %>%
  group_by(District) %>%
  summarise(Undervalued_Count = n(), .groups = "drop")

# Combine the data and calculate the percentage
apartment_district_analysis <- total_apartment_district %>%
  left_join(undervalued_apartments_district, by = "District") %>%
  mutate(
    Undervalued_Count = ifelse(is.na(Undervalued_Count), 0, Undervalued_Count), # Replace NAs with 0
    Percentage_Undervalued = (Undervalued_Count / Total) * 100
  ) %>%
  arrange(desc(Percentage_Undervalued))

# Display the results
print(apartment_district_analysis)

# Visualization of percentages by district
ggplot(apartment_district_analysis, aes(x = reorder(District, -Percentage_Undervalued), y = Percentage_Undervalued)) +
  geom_bar(stat = "identity", fill = "darkorange") +
  coord_flip() +
  ggtitle("Percentage of Undervalued Apartments by District") +
  xlab("District") +
  ylab("Percentage of Undervalued Apartments") +
  theme_minimal()
```

```{r}
undervalued_lands <- dataset_land_cleaned %>%
  filter(Price_Category_Cluster == "Undervalued")

# Count the total properties by district
total_land_district <- dataset_land_cleaned %>%
  group_by(District) %>%
  summarise(Total = n(), .groups = "drop")

# Count the undervalued properties by district
undervalued_land_district <- undervalued_lands %>%
  group_by(District) %>%
  summarise(Undervalued_Count = n(), .groups = "drop")

# Combine the data and calculate the percentage
land_district_analysis <- total_land_district %>%
  left_join(undervalued_land_district, by = "District") %>%
  mutate(
    Undervalued_Count = ifelse(is.na(Undervalued_Count), 0, Undervalued_Count), # Replace NAs with 0
    Percentage_Undervalued = (Undervalued_Count / Total) * 100
  ) %>%
  arrange(desc(Percentage_Undervalued))

# Display the results
print(land_district_analysis)

# Visualization of percentages by district
ggplot(land_district_analysis, aes(x = reorder(District, -Percentage_Undervalued), y = Percentage_Undervalued)) +
  geom_bar(stat = "identity", fill = "#33FFF5") +
  coord_flip() +
  ggtitle("Percentage of Undervalued Lands by District") +
  xlab("District") +
  ylab("Percentage of Undervalued Lands") +
  theme_minimal()
```

```{r}
undervalued_houses <- dataset_house_cleaned %>%
  filter(Price_Category_Cluster == "Undervalued")

# Count the total properties by district
total_house_district <- dataset_house_cleaned %>%
  group_by(District) %>%
  summarise(Total = n(), .groups = "drop")

# Count the undervalued properties by district
undervalued_house_district <- undervalued_houses %>%
  group_by(District) %>%
  summarise(Undervalued_Count = n(), .groups = "drop")

# Combine the data and calculate the percentage
house_district_analysis <- total_house_district %>%
  left_join(undervalued_house_district, by = "District") %>%
  mutate(
    Undervalued_Count = ifelse(is.na(Undervalued_Count), 0, Undervalued_Count), # Replace NAs with 0
    Percentage_Undervalued = (Undervalued_Count / Total) * 100
  ) %>%
  arrange(desc(Percentage_Undervalued))

# Display the results
print(house_district_analysis)

# Visualization of percentages by district
ggplot(house_district_analysis, aes(x = reorder(District, -Percentage_Undervalued), y = Percentage_Undervalued)) +
  geom_bar(stat = "identity", fill = "pink") +
  coord_flip() +
  ggtitle("Percentage of Undervalued Houses by District") +
  xlab("District") +
  ylab("Percentage of Undervalued Houses") +
  theme_minimal()
```

**- Apartments**

Districts such as Bragança, Guarda, Ilha de Porto Santo, and Portalegre
have 100% of their apartments classified as undervalued, while large
districts like Lisboa have a much lower proportion (23.2%).

**- Lands**

The districts with the highest proportion of undervalued lands are Viseu
(86.7%), Castelo Branco (84.2%), and Portalegre (81.8%). Districts such
as Lisboa and Faro show relatively low percentages.

**- Houses**

In the analysis of houses, the districts with the highest concentration
of undervalued properties are Viseu (100%), Guarda (90.3%), and Bragança
(90.1%). Lisboa and Faro, once again, show low percentages (15.0% and
14.0%, respectively).

**General Conclusion on Undervalued Properties**

**1. Districts with the highest concentration of undervalued
properties**

Districts with the highest proportions of undervalued properties are
consistently small or medium-sized in terms of the total number of
properties. For example:

-   Bragança, Guarda, and Portalegre rank among the districts with the
    highest percentage of undervalued properties across all three
    property categories (apartments, lands, and houses).

-   In particular, districts such as Bragança and Guarda show
    percentages of undervalued properties close to 100% in multiple
    categories, indicating that properties in these areas are
    significantly below the estimated market values.

**2. Districts with the lowest concentration of undervalued properties**

Large and economically active districts, such as:

-   Lisboa and Faro, consistently show a lower proportion of undervalued
    properties across all categories.

This suggests that properties in densely urbanized and tourist-heavy
areas tend to align more closely with their estimated market values,
likely due to higher competition and more accurate price assessments.

**3. Geographical patterns of undervaluation**

Rural or less populated districts:

-   Districts like Portalegre, Vila Real, and Castelo Branco tend to
    show high percentages of undervalued properties.

-   This could be due to lower demand in these areas, leading to market
    prices that do not fully reflect the properties’ potential value.

Urban or tourist regions:

-   Areas such as Lisboa, Setúbal, and Porto show lower percentages of
    undervalued properties, indicating tighter real estate markets
    aligned with economic dynamics.

**4. General implications**

Investment opportunities:

-   Districts like Bragança, Guarda, and Portalegre offer significant
    opportunities for investing in undervalued properties, as they show
    high percentages across all categories.

-   These areas may be ideal for buyers seeking properties with
    potential for future appreciation.

Consolidated markets:

-   Districts like Lisboa and Faro reflect more competitive markets,
    where properties are less likely to be undervalued. This may
    indicate higher economic development and stability in real estate
    prices.
